---
title: "How accurate is altitude from pressure geolocators?"
subtitle: "A global validation of GeoPressureR altitude against station barometers and radiosondes"
author: "Raphaël Nussbaumer"
date: today
format:
  gfm:
    toc: false
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 9
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: ".."
---

```{r setup}
library(data.table)
tab <- function(f) fread(file.path("output/tables", f))
hl <- tab("headline.csv")
h <- function(k, digits = 1) {
  v <- hl[metric == k, value]
  if (!length(v) || is.na(v)) return("NA")
  formatC(v, format = "f", digits = digits, big.mark = ",")
}
hi <- function(k) h(k, 0)
# Paths relative to report/, so the figures resolve both on GitHub (index.md) and in the HTML.
fig <- function(f) knitr::include_graphics(file.path("..", "output/figures", f), error = FALSE)
A <- tab("A_summary.csv")
B <- tab("B_height.csv")
gp_version <- readLines("output/tables/geopressurer_version.txt")
```

::: {.callout-note appearance="simple"}
## In short

**At ground level**, GeoPressureR's default (ERA5 single-levels) retrieves a geolocator's altitude
with a median per-site **bias of `r h("A_cons_abs_bias_median")` m** (90% of sites within
`r h("A_cons_abs_bias_p90")` m) and a **temporal scatter (SD) of `r h("A_cons_sd_median")` m**
(`r hi("A_cons_n_stations")` stations worldwide, 2022–2024). Pooled over all station-hours the mean
absolute error is **`r h("A_cons_mae")` m** and 95% of errors are below `r h("A_cons_p95")` m.
Terrain roughness is the main driver: errors roughly triple from flat to mountainous terrain.

**In flight**, altitude is **underestimated by about 1–1.5% of the height above the ground**
(`r h("B_rel_bias_1000_1500", 0)` m at 1–1.5 km, `r h("B_rel_bias_2000_3000", 0)` m at 2–3 km,
`r h("B_rel_bias_5000_6000", 0)` m at 5–6 km), because the formula assumes a standard temperature
profile. The error is largest in continental and polar winters (surface inversions). Averaged over
the heights at which tracked birds fly, flight adds a mean absolute error of
**`r h("B_bird_weighted_mae")` m**.

**ERA5-Land** is much worse for absolute altitude (MAE `r h("A_cons_land_mae")` m) and should not be
used for it.
:::

# Why this study

GeoPressureR (and GeoPressureAPI behind it) converts the pressure a geolocator records into an
altitude using ERA5 reanalysis: the surface pressure $p_0$ and 2 m temperature $T_0$ at the bird's
location and hour, the ERA5 orography $z_0$, and the barometric formula with a standard lapse rate
$L = -6.5$ K/km,

$$
z = z_0 + \frac{T_0}{L}\left[\left(\frac{p}{p_0}\right)^{-R L / (g M)} - 1\right].
$$

Users want to know how good that altitude is, where and when it is worse, and why. An earlier check
over the Alps
([GeoPressureAPI#27](https://github.com/GeoPressure/GeoPressureAPI/issues/27)) found ~9 m mean
absolute error with ERA5 single-levels and led to making single-levels the default. This study extends
that check to the whole globe, a full annual cycle, three recent years, and — using radiosondes —
to the heights at which birds fly.

# Methods

## What is validated

The exact GeoPressureR code path is evaluated: grid-cell snapping, nearest-hour matching, the ERA5
orography files and `pressure_to_altitude()` are taken from GeoPressureR `r gp_version`, reading
ERA5 from the ECMWF ARCO archive (`pressurepath_create(source = "arco")`). A cross-check against
GeoPressureAPI (Earth Engine) on `r if (file.exists("output/tables/api_crosscheck_summary.csv")) tab("api_crosscheck_summary.csv")$n else "NA"`
station-hours found a maximum difference of `r h("api_max_abs_diff", 2)` m, so every result below
applies to both entry points.

Two ERA5 products are compared: **ERA5 single-levels** (0.25°, the default) and **ERA5-Land**
(0.1°). The mixed `"both"` option is not evaluated separately: over land it is ERA5-Land.

## Reference data

**Tier A — surface barometers (ground level).** HadISD v3.4.3.2025f (Dunn et al. 2012, 2016), a
quality-controlled subset of NOAA ISD. Its station-level pressure is treated as a geolocator reading
and the retrieved altitude is compared with the station elevation. Stations were chosen for spatial
balance: in every 5°×5° cell the highest, the lowest and one random station, plus every station
above 1000 m. Stations had to report station pressure on at least half the days of each of
2022–2024, at least 6-hourly; the set was then thinned to one random station per 5° cell plus one
per 2.5° cell above 1000 m. ERA5-Land was evaluated on a random 200 of these stations, and a random
100 reporting in 1990, 2005 and 2015 test for changes over time.

**Tier B — radiosondes (0–6 km above ground).** IGRA2 (Durre et al. 2006, 2018), 2023–2024. At
every level the sonde reports pressure and geopotential height; the pressure is given to the
altitude formula at the launch site and hour of that level, and the retrieved altitude is compared
with the sonde's height. Stations: the highest and one random active station per 10°×10° cell, plus
all above 1000 m.

## Separating accuracy from precision, and handling errors in the reference

At ~5–10 m, errors in the reference itself matter. Three choices keep them out of the headline
numbers:

- **Accuracy vs precision.** Each station's error is split into its mean (the *bias*: constant in
  time, relevant for absolute altitude) and the SD around that mean (*precision*: relevant for
  altitude changes). A wrong station elevation shifts only the bias, so precision is robust to
  metadata errors.
- **Trusted metadata subset.** Accuracy is also reported for stations whose HadISD elevation agrees
  within 20 m with an independent DEM (Mapzen terrain tiles via OpenTopoData).
- **Height relative to the surface level (Tier B).** The error at a level minus the error at the same
  sounding's surface level cancels the station elevation entirely, isolating how the error grows
  with height.

Gross observation errors were flagged when more than 100 m (Tier A) or 150 m (Tier B) and 10 robust
SD from the station's (or height bin's) median error: `r h("A_gross_fraction", 5)` of Tier A and
`r h("B_gross_fraction", 5)` of Tier B observations.

Pooled statistics give every station equal weight.

**Reference consistency screen.** 
`r hi("A_n_inconsistent")` of `r hi("A_n_stations")` Tier A stations (5%) carry a constant offset
larger than any plausible extrapolation error (|bias| > 30 m + 10% of the gap between station
elevation and ERA5 orography, equivalent to a ~28 K error in the assumed layer temperature). Their
pressure tracks ERA5 hour by hour (median SD ~5 m) but at a fixed offset of typically 75 m: the
signature of a pressure reported at a different height than the listed elevation (datum or
metadata error), not of a reanalysis error. They are included in the "all stations" row and
excluded elsewhere. A few could be genuine errors in deep valleys; this is a limitation.

## Drivers

Candidate drivers of accuracy and precision: the difference between station elevation and ERA5
orography, sub-grid terrain roughness (ERA5 `sdor`), latitude, Köppen-Geiger climate zone (station
level); local solar hour, season, boundary layer height, skin − 2 m temperature (a proxy for
surface inversions) and 6 h surface pressure tendency (observation level). Their effect is
estimated with GAMs (mgcv) and ranked by the deviance explained lost when each is dropped.

For Tier B, the error expected from the temperature assumption alone is
$\Delta z \cdot (\bar T_\text{assumed} - \bar T_\text{obs}) / \bar T_\text{obs}$, where
$\bar T_\text{obs}$ is the sonde's log-pressure-weighted mean temperature of the layer below the level.

# Results: ground level (Tier A)

```{r}
knitr::kable(
  A[group == "all", .(subset, dataset = fifelse(dataset == "sl", "single-levels", "land"),
    stations = n_stations, `obs` = format(n_obs, big.mark = ","),
    `median |bias|` = round(abs_bias_median, 1), `p90 |bias|` = round(abs_bias_p90, 1),
    `median SD` = round(sd_median, 1), `p90 SD` = round(sd_p90, 1), MAE = round(mae, 1),
    RMSE = round(rmse, 1), `p95 |err|` = round(p95_abs, 1))],
  caption = "Altitude error at ground level, 2022-2024 (m)."
)
```

```{r}
#| fig-cap: "Accuracy: mean error per station."
fig("A_map_bias.png")
```

```{r}
#| fig-cap: "Precision: temporal SD of the error per station."
fig("A_map_sd.png")
```

```{r}
#| fig-cap: "Share of stations below a given |bias|, ERA5 single-levels vs ERA5-Land."
#| out-width: 70%
fig("A_sl_vs_land.png")
```

## By terrain, elevation, latitude and climate

```{r}
knitr::kable(
  A[group != "all" & dataset == "sl", .(group, subset, stations = n_stations,
    `median |bias|` = round(abs_bias_median, 1), `p90 |bias|` = round(abs_bias_p90, 1),
    `median SD` = round(sd_median, 1), `p90 SD` = round(sd_p90, 1), MAE = round(mae, 1),
    `p95 |err|` = round(p95_abs, 1))],
  caption = "ERA5 single-levels, by station class (m). Terrain = sub-grid orography SD in the ERA5 cell."
)
```

## Drivers

```{r}
#| fig-cap: "Left: bias against the gap between station elevation and ERA5 orography. Right: precision against sub-grid terrain roughness."
fig("A_drivers.png")
```

```{r}
d <- tab("A_drivers_station.csv")
knitr::kable(d[, .(response, term, `deviance explained (full)` = round(dev_expl_full, 3),
  `lost when dropped` = round(dev_expl_drop, 3), n)], caption = "Station-level drivers (GAM).")
d <- tab("A_drivers_obs.csv")
knitr::kable(d[, .(term, `deviance explained (full)` = round(dev_expl_full, 3),
  `lost when dropped` = round(dev_expl_drop, 3), n)],
  caption = "Observation-level drivers of the de-biased absolute error (GAM with station random effect).")
```

```{r}
#| fig-cap: "Diurnal and seasonal cycle of the de-biased error (RMS), by climate zone."
fig("A_cycles.png")
```

```{r}
#| fig-cap: "Change over time, for stations reporting in every year."
#| out-width: 70%
if (file.exists("output/figures/A_era.png")) fig("A_era.png")
```

```{r}
#| results: asis
if (file.exists("output/tables/A_era.csv")) {
  e <- tab("A_era.csv")
  print(knitr::kable(e[, lapply(.SD, function(x) round(x, 1)), .SDcols = names(e)],
    caption = "Change over time for stations reporting in every year (m)."))
}
```

# Results: in flight (Tier B)

```{r}
knitr::kable(
  B[, .(method, `height above ground` = hbin, n = format(n, big.mark = ","),
    bias = round(bias, 1), SD = round(sd, 1), MAE = round(mae, 1), RMSE = round(rmse, 1),
    `p95 |err|` = round(p95_abs, 1))],
  caption = "Error relative to the surface level of the same sounding, by height (m)."
)
```

```{r}
#| fig-cap: "Bias and SD of the altitude error against height above ground, for GeoPressureR and alternative formulations, with the height distribution of geolocator flight points (right)."
fig("B_height.png")
```

```{r}
#| fig-cap: "The height-dependent error is almost entirely the difference between the temperature profile the formula assumes and the one the sonde observed."
#| out-width: 70%
fig("B_temperature.png")
```

```{r}
#| fig-cap: "Height-dependent bias by climate zone and season (solid: summer half-year, dashed: winter half-year; southern hemisphere shifted by six months)."
#| out-width: 80%
fig("B_height_climate.png")
```

```{r}
bw <- tab("B_bird_weighted.csv")
knitr::kable(bw[, .(method, bias = round(bias, 1), MAE = round(mae, 1), RMSE = round(rmse, 1))],
  caption = "Error averaged over the height distribution of geolocator flight points (m).")
```

```{r}
d <- tab("B_height_daynight_season.csv")[method == "GeoPressureR (ERA5 single-levels)"]
knitr::kable(dcast(d, hbin ~ daynight + season, value.var = "bias", fun.aggregate = function(x)
  round(x, 1)), caption = "Bias by height, day/night and season (m).")
```

# Discussion and guidance for users

```{r}
tr <- A[dataset == "sl"]
t <- function(sub, col) round(tr[subset == sub][[col]], 1)
e <- if (file.exists("output/tables/A_era.csv")) tab("A_era.csv") else NULL
te <- tab("B_temperature_explanation.csv")
bwv <- tab("B_bird_weighted.csv")
bws <- function(m, col) round(bwv[method == m][[col]], 1)
```

## What users can expect

| Situation | Typical error | Driver |
|---|---|---|
| Absolute altitude on the ground, flat terrain | bias `r t("flat (<20 m)", "abs_bias_median")` m (p90 `r t("flat (<20 m)", "abs_bias_p90")` m), MAE `r t("flat (<20 m)", "mae")` m | ERA5 surface pressure at 0.25° |
| Absolute altitude on the ground, mountains (sub-grid SD > 300 m) | bias `r t("mountain (>300 m)", "abs_bias_median")` m (p90 `r t("mountain (>300 m)", "abs_bias_p90")` m), MAE `r t("mountain (>300 m)", "mae")` m | terrain the 0.25° grid cannot resolve |
| Altitude *changes* at one site (precision) | SD `r t("flat (<20 m)", "sd_median")` m (flat) to `r t("mountain (>300 m)", "sd_median")` m (mountains) | as above; almost no diurnal or seasonal cycle |
| In flight, 1–1.5 km above ground | `r h("B_rel_bias_1000_1500", 0)` m bias, RMSE `r h("B_rel_rmse_1000_1500", 0)` m (added to the ground-level error) | assumed temperature profile |
| In flight, 2–3 km above ground | `r h("B_rel_bias_2000_3000", 0)` m bias, RMSE `r h("B_rel_rmse_2000_3000", 0)` m | as above; ~2x in continental/polar winter |

**On the ground, the error is mostly a fixed offset per place, not noise.** Station identity
explains almost all of the explainable variation in the de-biased error; hour of day, season,
boundary layer height, surface inversions and pressure tendency each add less than 0.5% of
deviance. Precision (SD ~4 m) is therefore better than absolute accuracy (MAE ~7 m), and roughly
constant in time. What changes the error is *where*: the gap between the
true ground and ERA5's smoothed orography, and sub-grid terrain roughness.

**In flight, the error grows with height and is systematic.** The barometric formula extrapolates
the ERA5 2 m temperature upwards with a standard lapse rate of −6.5 K/km and ignores humidity. The
real atmosphere is usually warmer than that (shallower lapse rates, surface inversions, and the
virtual-temperature effect of water vapour), so the layer is too cold in the formula and the bird is
placed too low. The error predicted from the sonde's own temperature profile explains
`r round(100 * te$r2)`% of the variance (r = `r round(te$r, 2)`, slope `r round(te$slope, 2)`); the
remaining offset of `r round(te$intercept)` m is consistent with humidity, which the dry-bulb
comparison leaves out. It is worst over cold continental and polar surfaces in winter, where strong
inversions sit under much warmer air, and in the humid tropics.

**Flight altitudes from GeoPressureR are therefore conservative (slightly too low).** For a bird at
2 km above the ground, expect it about 25–35 m too low, with a scatter (SD) of 30–50 m; roughly
double that in a continental or polar winter.

**ERA5 has improved.** For the same stations, the median ground-level bias fell from
`r if (!is.null(e)) round(e[year == min(year), abs_bias_median], 1) else "NA"` m in
`r if (!is.null(e)) min(e$year) else "NA"` to
`r if (!is.null(e)) round(e[year == max(year), abs_bias_median], 1) else "NA"` m in
`r if (!is.null(e)) max(e$year) else "NA"`, and the precision improved by about a third. Tracks
from the 1990s and earlier should be expected to be somewhat less accurate.

**ERA5-Land vs ERA5 single-levels.** ERA5-Land's median ground-level bias is about four times
larger. Its height-*relative* errors in flight are similar to single-levels, because the error
relative to the surface level cancels its static orography inconsistency. This confirms that its
problem is a fixed spatial offset
([GeoPressureAPI#27](https://github.com/GeoPressure/GeoPressureAPI/issues/27)), which matters
for absolute altitude and for comparing altitudes between places.

## Could the formula do better?

Two simple changes remove most of the average bias in flight: using the 2 m virtual temperature
(from ERA5 2 m dewpoint), and a −5 K/km lapse rate. Averaged over bird flight heights, the bias drops
from `r bws("GeoPressureR (ERA5 single-levels)", "bias")` m to
`r bws("virtual temperature + lapse -5 K/km", "bias")` m and the RMSE from
`r bws("GeoPressureR (ERA5 single-levels)", "rmse")` m to
`r bws("virtual temperature + lapse -5 K/km", "rmse")` m. The scatter barely changes, though. It
comes from day-to-day and seasonal departures from any fixed profile, mainly inversions. Removing it
needs the actual temperature profile, i.e. ERA5 pressure-level temperature or geopotential. That
would be worth evaluating as a GeoPressureR improvement for in-flight altitude; it is not what
GeoPressureR does today, and nothing here changes its ground-level behaviour.

## Limitations

- **Not independent of ERA5.** Station surface pressure and radiosondes are both assimilated by
  ERA5, so errors *at these sites* are optimistic. Remote areas without observations are likely
  somewhat worse, especially for ground-level bias.
- **The sensor is not included.** Geolocator pressure sensors add their own offset, drift and
  resolution (1 hPa ≈ 8 m near the ground). The numbers here are the error of the retrieval method
  given a perfect pressure reading.
- **Radiosonde geometry.** The ERA5 column at the launch site is used; balloon drift (typically a few
  km below 5 km) is ignored. Sonde heights are themselves derived hypsometrically. The ~10 m scatter
  within the first 100 m partly reflects integer-metre reporting and interpolated low levels.
- **Reference screen.** 5% of surface stations were excluded as datum/metadata errors. A few may be
  genuine valley errors, so the mountain numbers may be slightly optimistic.
- **Sampling.** Stations are thinned for spatial balance, but coverage is still sparse in parts of
  Africa, South America and over the oceans (Tier A is land-only).

# Data and code availability

Code: <https://github.com/GeoPressure/altitude-validation>. ERA5 and ERA5-Land: Copernicus Climate
Change Service (Hersbach et al. 2020; Muñoz-Sabater et al. 2021), via the ECMWF ARCO archive.
HadISD: Met Office Hadley Centre, Non-Commercial Government Licence; this product may contain data
governed by WMO Resolution 40 Annex 1. IGRA2: NOAA NCEI. Bird heights: GeoLocator master data
package (<https://doi.org/10.5281/zenodo.18187092>).
