---
title: "How accurate is altitude from pressure geolocators?"
subtitle: "A global validation of GeoPressureR altitude against station barometers and radiosondes"
author: "Raphaël Nussbaumer"
date: today
format:
  gfm:
    toc: false
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 9
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: ".."
---

```{r setup}
invisible(Sys.setlocale("LC_CTYPE", "en_US.UTF-8"))
library(data.table)
options(knitr.kable.NA = "")
tab <- function(f) fread(file.path("output/tables", f))
hl <- tab("headline.csv")
h <- function(k, digits = 1) {
  v <- hl[metric == k, value]
  if (!length(v) || is.na(v)) return("NA")
  formatC(v, format = "f", digits = digits, big.mark = ",")
}
hi <- function(k) h(k, 0)
hv <- function(k) hl[metric == k, value]
r1 <- function(x) formatC(x, format = "f", digits = 1)
pct <- function(x) paste0(round(100 * x), "%")
pct2 <- function(k) sprintf("%.2f%%", 100 * hv(k))
nm <- function(x) paste(tools::toTitleCase(tolower(x)), collapse = ", ")
# Paths relative to report/, so the figures resolve both on GitHub (index.md) and in the HTML.
fig <- function(f) knitr::include_graphics(file.path("..", "output/figures", f), error = FALSE)
gp_version <- readLines("output/tables/geopressurer_version.txt")

# Tier A
A <- tab("A_summary.csv")
lab_ref <- "reference set (plausible offset, no step change)"
Asl <- A[dataset == "sl"]
t <- function(sub, col) r1(Asl[subset == sub][[col]])
sa <- tab("A_station_stats.csv")
ex <- tab("A_excluded.csv")
nbp <- tab("A_neighbours.csv")
bf <- tab("A_screen_basis_flat.csv")
bg <- tab("A_screen_basis_gap.csv")
sens <- tab("A_sensitivity.csv")
era <- if (file.exists("output/tables/A_era.csv")) tab("A_era.csv") else NULL
api <- tab("api_crosscheck_summary.csv")
ref <- sa[reference == TRUE]
imp <- sa[consistent == FALSE]
ex_nb <- ex[is.finite(nb_bias_median)]
ref_nb <- nbp[id %in% ref$id & abs(bias) > 15, .(bias = bias[1], nb = median(nb_bias)), by = id]
n_shared <- ref_nb[sign(nb) == sign(bias) & abs(nb) >= 0.5 * abs(bias), .N]

# Tier B
B <- tab("B_height.csv")
Bsl <- B[method == "GeoPressureR (ERA5 single-levels)"]
bu <- tab("B_stations_used.csv")
st <- tab("stations.csv")
stB <- st[tier == "B" & id %in% bu$id]
te <- tab("B_temperature_explanation.csv")

# Formula variants (07b_formula.R)
ff <- tab("formula_fit.csv")
fcv <- tab("formula_cv.csv")
fbw <- tab("formula_bird_weighted.csv")
fh <- tab("formula_height.csv")
fg <- tab("formula_ground.csv")
fgrp <- tab("formula_lapse_by_group.csv")
ftv <- tab("formula_tv_correction.csv")
L_tv <- ff[temperature == "Tv", L_fit]
L_t2m <- ff[temperature == "T2m", L_fit]
fc <- function(k, col, digits = 1) formatC(fcv[code == k][[col]], format = "f", digits = digits)
fhb <- function(k, bin) round(fh[code == k & hbin == bin, bias])
```

::: {.callout-note appearance="simple"}
## In short

- **Ground level.** GeoPressureR's default (ERA5 single-levels) places a geolocator on the ground
  with a median per-site **bias of `r h("A_cons_abs_bias_median")` m** (90% of sites within
  `r h("A_cons_abs_bias_p90")` m) and a **temporal scatter (SD) of `r h("A_cons_sd_median")` m**.
  Pooled MAE: **`r h("A_cons_mae")` m** (`r hi("A_cons_n_stations")` stations, 2022–2024).
- **On the ground, terrain drives the error.** It is a fixed offset per place, and it roughly
  triples from flat terrain to mountains.
- **In flight, altitude is too low by ~1–1.5% of the height above ground**:
  `r h("B_rel_bias_1000_1500", 0)` m at 1–1.5 km, `r h("B_rel_bias_2000_3000", 0)` m at 2–3 km. The
  cause is the formula's standard temperature profile. Over the heights birds fly, flight adds a MAE
  of **`r h("B_bird_weighted_mae")` m**.
- **ERA5-Land is much worse** for absolute altitude (MAE `r h("A_cons_land_mae")` m). Don't use it
  for altitude.
- **The formula can be improved.** Virtual temperature plus a fitted lapse rate of
  `r sprintf("%.1f", L_tv)` K/km cuts the in-flight bias from `r fc("current", "bias_mean")` m to
  `r fc("tv_lapse", "bias_mean")` m on held-out stations, with no change on the ground. GeoPressureR
  does not do this (yet).
:::

# Introduction

**Pressure geolocators record atmospheric pressure, not altitude.** Altitude is derived
afterwards, by comparing the recorded pressure with a weather reanalysis. GeoPressureR (and
GeoPressureAPI behind it) does this with ERA5. The altitude is then used to:

- tell flight from rest, and reconstruct flight altitude and climbs;
- compare the altitude of resting sites with the terrain, which helps with positioning.

Users need to know how good that altitude is, where and when it is worse, and why.

## How GeoPressureR computes altitude

For a pressure $p$ at a given place and hour, GeoPressureR reads three ERA5 fields at the nearest
grid cell and hour: surface pressure $p_0$, 2 m temperature $T_0$ and orography $z_0$. It then
applies the barometric formula with the standard lapse rate $L = -6.5$ K/km:

$$
z = z_0 + \frac{T_0}{L}\left[\left(\frac{p}{p_0}\right)^{-R L / (g M)} - 1\right].
$$

This is the hypsometric equation (Wallace & Hobbs 2006, ch. 3) for one assumed temperature
profile:

$$
z - z_0 = \frac{R_d}{g}\,\bar T_v \,\ln\frac{p_0}{p},
$$

where $\bar T_v$ is the (log-pressure weighted) mean *virtual* temperature of the layer between
$p_0$ and $p$. The assumed profile:

- starts at the ERA5 2 m temperature and decreases at −6.5 K/km, as in the ICAO standard atmosphere
  (ICAO 1993; ISO 1975);
- treats the air as dry ($T_v = T$).

## Where the error comes from

- **ERA5 at the surface.** $p_0$ and $z_0$ are 0.25° (~28 km) averages. Terrain the grid can't
  resolve puts the real ground above or below $z_0$. This mostly matters on the ground.
- **The assumed temperature profile.** The layer thickness is proportional to $\bar T_v$, so an
  error in the assumed mean temperature is the same *relative* error in height (below). This
  mostly matters in flight.
- **The sensor.** A geolocator's own offset, drift and resolution (1 hPa ≈ 8 m) come on top. They
  are not evaluated here.

$$
\frac{\delta (z - z_0)}{z - z_0} = \frac{\bar T_\text{assumed} - \bar T_{v,\text{true}}}{\bar T_{v,\text{true}}}
$$

An assumed layer 3 K (≈1%) too cold puts a bird at 2 km about 20 m too low. On the ground, $z - z_0$
is only the small gap to the ERA5 orography (median `r round(fg$median_abs_dz)` m), so the
assumption barely matters there.

## Alternatives to the standard profile

The assumed layer is usually too cold, for two separate reasons:

1. **Humidity.** Moist air is lighter, so a moist layer is thicker. The hypsometric equation
   requires the virtual temperature $T_v = T\,(1 + 0.608\,q)$, with $q$ the specific humidity.
   Using $T$ instead is a simplification, not a modelling choice, and correcting it has no free
   parameter.
2. **Lapse rate.** −6.5 K/km is a mean over the whole troposphere (0–11 km). Near the surface, lapse
   rates vary with latitude and season (Stone & Carlson 1979; Rolland 2003) and are often
   shallower, e.g. 3.9–5.2 K/km annual means in the Cascade Mountains (Minder et al. 2010). Surface
   inversions make them shallower still, or reverse them.

Both are simple changes to the formula. This report tests whether they help.

## Aims

An earlier check over the Alps
([GeoPressureAPI#27](https://github.com/GeoPressure/GeoPressureAPI/issues/27)) found ~9 m MAE with
ERA5 single-levels and made single-levels the default. This study extends it to:

1. **the globe, the annual cycle and three recent years**: accuracy and precision on the ground;
2. **the heights birds fly (0–6 km)**, using radiosondes;
3. **the drivers**: terrain, climate, season, time of day, and ERA5's change over time;
4. **ERA5-Land vs ERA5 single-levels**;
5. **two formula improvements**: virtual temperature, and a lapse rate fitted to the radiosondes.

# Methods

## Altitude retrieval evaluated

- **The exact GeoPressureR code path.** Grid snapping, nearest-hour matching, the ERA5 orography and
  `pressure_to_altitude()` all come from GeoPressureR `r gp_version`. ERA5 is read from the ECMWF
  ARCO archive (`pressurepath_create(source = "arco")`).
- **GeoPressureAPI gives the same result.** Across `r format(api$n, big.mark = ",")` station-hours
  at `r api$stations` stations, the API (Earth Engine) and ARCO differ by at most
  `r h("api_max_abs_diff", 2)` m. All results apply to both entry points.
- **Two ERA5 products.** ERA5 single-levels (0.25°, the default; Hersbach et al. 2020) and ERA5-Land
  (0.1°; Muñoz-Sabater et al. 2021). GeoPressureR's `"both"` option is ERA5-Land over land, so it is
  not evaluated separately.

## Data

```{r}
#| label: tbl-data
#| tbl-cap: "Reference data."
dt <- data.table(
  ` ` = c("Source", "Years", "Stations selected", "Stations used", "Observations",
    "Station elevation (median, range)", "Stations above 1000 m", "Reference height"),
  `Tier A: surface barometers` = c("HadISD v3.4.3.2025f", "2022–2024 (+ 1990, 2005, 2015 for 100)",
    hi("A_n_candidates"), paste0(hi("A_n_stations"), " (", hi("A_cons_n_stations"),
      " in the reference set)"),
    paste(h("A_n_obs", 0), "station-hours"),
    sprintf("%d m (%d to %d m)", round(median(sa$elev)), round(min(sa$elev)), round(max(sa$elev))),
    pct(mean(sa$elev > 1000)), "station elevation"),
  `Tier B: radiosondes` = c("IGRA v2.2", "2023–2024", hi("B_n_candidates"), hi("B_n_stations"),
    paste(h("B_n_soundings", 0), "soundings,", h("B_n_levels", 0), "levels"),
    sprintf("%d m (%d to %d m)", round(median(stB$elev)), round(min(stB$elev)),
      round(max(stB$elev))),
    pct(mean(stB$elev > 1000)), "sonde geopotential height")
)
knitr::kable(dt)
```

```{r}
#| label: fig-stations
#| fig-cap: "Stations used. Right: station elevations (log scale)."
fig("data_stations.png")
```

### Tier A: surface barometers (HadISD)

**Source.** HadISD (Dunn et al. 2012, 2016) is a quality-controlled subset of NOAA's Integrated
Surface Database (ISD; Smith et al. 2011). Its **station-level pressure** is treated as a
geolocator reading, and the altitude retrieved from it is compared with the station elevation.

**Selection.** The aim is spatial balance, not network density:

- **Candidates** (`r hi("A_n_candidates")`): in every 5°×5° cell, the highest, the lowest and one
  random station, plus every station above 1000 m.
- **Coverage** (`r hi("A_n_coverage")` kept): station pressure on at least half the days of each of
  2022–2024, at least 6-hourly on average.
- **Thinning** (`r hi("A_n_stations")` kept): one random station per 5° cell, plus one per 2.5° cell
  above 1000 m. This keeps mountains well represented.
- **Subsets.** ERA5-Land is read for a random 200 stations. A random 100 that also report in 1990,
  2005 and 2015 are used to test change over time.

**Gross errors.** Single observations more than 100 m and 10 robust SD from their station's median
error are dropped (`r pct2("A_gross_fraction")` of observations).

**Reference screen.** At ~5–10 m, errors in the reference itself matter. They are well documented:

- SYNOP surface pressure is biased at several hundred stations, often by several hPa. The cause is
  "mostly … incorrect assumptions about the station heights", and the bias stays fairly constant in
  time (Vasiljevic et al. 2006).
- ECMWF's monitoring guidance lists the usual causes (ECMWF 2025): wrong lat/lon/elevation metadata,
  GNSS elevations without the geoid correction, sensor or encoding errors, and step jumps when the
  barometer height or position changes.
- WMO defines the datum of station pressure ($H_p$) separately from the ground elevation (WMO 2023),
  so the two can legitimately differ.
- HadISD is quality-controlled but **not homogenised**, and it merges records of co-located stations
  (Dunn et al. 2012, 2016).
- ERA5 bias-corrects or rejects such stations when it assimilates them (Vasiljevic et al. 2006;
  Hersbach et al. 2020). Where one station and ERA5 disagree by a constant offset, the station is
  usually at fault.

Two rules therefore remove stations from the **reference set** (`r hi("A_n_excluded")` of
`r hi("A_n_stations")`, `r pct(hv("A_n_excluded") / hv("A_n_stations"))`):

1. **Implausible offset** (`r hi("A_n_inconsistent")`): |bias| > 30 m + 10% of the gap between
   station elevation and ERA5 orography. That is about twice what ERA5 can produce, judged from
   data that don't use the station's own bias:
   - where ERA5 needs no extrapolation (flat, gap < 30 m), 95% of stations are within
     `r round(bf$p95)` m;
   - over a height gap, radiosondes put the extrapolation error at `r r1(max(bg$median_pct))` to
     `r r1(min(bg$median_pct))`% of the gap, and within `r round(max(bg$p99_abs_pct))`% for 99% of
     stations.
2. **Step change** (`r hi("A_n_step")`): a clear jump in the station's error during 2022–2024. Each
   station's monthly median errors are fitted with a month-of-year effect plus one step, and a
   station is flagged if |step| ≥ 10 m and |t| ≥ 10. ERA5 doesn't jump at one site; a station does
   when it moves or changes barometer. Examples:
   `r nm(head(ex[reason == "step change"][order(-abs(step))]$name, 3))`
   (`r paste(head(ex[reason == "step change"][order(-abs(step))]$step, 3), collapse = ", ")` m).

**Independent checks confirm that the screen removes station errors, not ERA5 errors:**

- **Neighbours.** `r nrow(ex_nb)` excluded stations have another HadISD station within 50 km. ERA5
  fits every one of those neighbours to within `r ceiling(max(abs(ex_nb$nb_bias_median)))` m, while
  the excluded stations are off by up to `r format(max(abs(ex_nb$bias)), big.mark = ",")` m.
  Conversely, `r n_shared` of `r nrow(ref_nb)` tested reference stations with |bias| > 15 m share
  their offset with a neighbour. Those are real ERA5 errors, mostly in mountains, and they are kept.
- **No mechanism in ERA5.** `r sum(imp$dem_refutes == FALSE, na.rm = TRUE)` of the
  `r nrow(imp)` implausible offsets are at stations whose elevation the DEM confirms, often in flat
  terrain next to ERA5's orography (e.g. `r nm(head(imp[dem_refutes == FALSE & abs(dz_sl) < 30][order(-abs(bias_sl))]$name, 3))`).
- **Constant in time.** Excluded stations track ERA5 hour by hour (median SD
  `r r1(median(imp$sd_sl))` m) at a fixed offset, the signature of a station-height error.

**No DEM filter.** A DEM check (SRTM, ASTER) mostly measures coordinate precision:
`r pct(mean(sa$minute_coord, na.rm = TRUE))` of coordinates are rounded to whole arc-minutes
(±900 m). At `r hi("A_n_dem_refutes_reference")` good stations, the DEM contradicts an elevation
that ERA5 reproduces. The DEM is used only as supporting evidence (Appendix A).

**What remains.** The reference set still holds small reference errors no screen can catch: steps
< 10 m, barometers a few metres above the listed ground, rounded elevations. Its accuracy is
therefore an upper bound on ERA5's. The median bias is robust to the screen
(`r r1(min(sens$abs_bias_median))`–`r r1(max(sens$abs_bias_median))` m across all variants,
Appendix A); MAE and RMSE are not.

### Tier B: radiosondes (IGRA2)

**Source.** IGRA v2.2 (Durre et al. 2006, 2018). At every level a sonde reports pressure and
geopotential height. The pressure is given to the altitude formula at the launch site and the hour
of that level, and the result is compared with the sonde's height.

**Selection and filtering.**

- **Candidates** (`r hi("B_n_candidates")`): in every 10°×10° cell, the highest and one random
  active station, plus every station above 1000 m. `r hi("B_n_with_data")` have data in 2023–2024.
- **Levels**: all levels up to 6 km above ground, the upper end of bird flight. The time of each
  level comes from the reported elapsed time, or assumes a 5 m/s ascent.
- **Enough soundings**: stations weigh equally, so each needs ≥100 soundings
  (`r hi("B_n_stations")` kept).
- **Gross errors**: levels more than 150 m and 10 robust SD from the median of their height bin are
  dropped (`r pct2("B_gross_fraction")` of levels).
- **No station screen needed**: errors are taken relative to the sounding's surface level (below),
  which cancels the station elevation.

### Bird flight heights

The heights at which tracked birds fly are used to weight the in-flight errors. They come from the
flight points of the GeoLocator master data package (<https://doi.org/10.5281/zenodo.18187092>),
binned by height above ground.

## Error metrics

- **Accuracy vs precision.** Each station's error splits into its mean (the **bias**: constant in
  time, matters for absolute altitude) and the SD around it (**precision**: matters for altitude
  changes). A wrong station elevation shifts only the bias.
- **Height relative to the surface (Tier B).** Error at a level minus error at the same sounding's
  surface level. This isolates how the error grows with height.
- **Equal station weights.** Pooled statistics weight stations equally, so dense networks don't
  dominate.
- **Bird-weighted error.** Tier B height bins weighted by the share of bird flight points in them.

## Drivers

- **Candidates.** Station level: gap between station elevation and ERA5 orography, sub-grid terrain
  roughness (ERA5 `sdor`), latitude, Köppen-Geiger climate zone. Observation level: local solar
  hour, season, boundary layer height, skin − 2 m temperature (a proxy for surface inversions), 6 h
  surface pressure tendency.
- **Model.** GAMs (mgcv). Each driver is ranked by the deviance explained lost when it is dropped.
- **Temperature explanation (Tier B).** The error expected from the temperature assumption alone is
  $\Delta z \cdot (\bar T_\text{assumed} - \bar T_\text{obs}) / \bar T_\text{obs}$, with
  $\bar T_\text{obs}$ the sonde's log-pressure weighted mean temperature below the level.

## Formula variants

- **Virtual temperature.** $q$ comes from the ERA5 2 m dewpoint and surface pressure (vapour
  pressure from Bolton 1980). Surface humidity is assumed to hold through the layer, as in pressure
  reduction to sea level (WMO 1968).
- **Fitted lapse rate.** $L$ minimises the bird-weighted squared error relative to the surface
  (Tier B). It is fitted both with $T_0$ and with $T_{v,0}$.
- **Robustness.** 20 random 50/50 station splits (fit on one half, evaluate on the other); separate
  fits per climate zone and half-year; an alternative objective that weights all height bins
  equally. The effect on the ground is evaluated on Tier A.
- **Headline numbers stay those of the current formula.** Variants appear only where labelled.

# Results

## Ground level

```{r}
#| label: tbl-ground
#| tbl-cap: "Altitude error at ground level, 2022–2024 (m). |bias|: per-station mean error. SD: temporal scatter per station. MAE, RMSE, p95: pooled over station-hours, stations weighted equally."
g <- A[group == "all" & (dataset == "sl" | subset == lab_ref)]
g[, row := fifelse(dataset == "land", "reference set, ERA5-Land",
  fifelse(subset == lab_ref, "reference set", "all stations"))]
knitr::kable(g[order(dataset != "sl", subset != "all stations"), .(` ` = row,
  stations = n_stations, `median |bias|` = r1(abs_bias_median), `p90 |bias|` = r1(abs_bias_p90),
  `median SD` = r1(sd_median), MAE = r1(mae), RMSE = r1(rmse), `p95 |err|` = r1(p95_abs))])
```

- **Accuracy: median |bias| `r h("A_cons_abs_bias_median")` m, 90% of sites within
  `r h("A_cons_abs_bias_p90")` m.** Biases of both signs occur everywhere; the largest are in
  mountains (@fig-maps).
- **Precision: SD `r h("A_cons_sd_median")` m.** Precision is better than accuracy: at a given place,
  the error is mostly a fixed offset.
- **Terrain drives both** (@fig-class). From flat to mountainous terrain:
  - median |bias| `r t("flat (<20 m)", "abs_bias_median")` → `r t("mountain (>300 m)", "abs_bias_median")` m;
  - MAE `r t("flat (<20 m)", "mae")` → `r t("mountain (>300 m)", "mae")` m;
  - SD `r t("flat (<20 m)", "sd_median")` → `r t("mountain (>300 m)", "sd_median")` m.

  Latitude and climate matter much less.

```{r}
#| label: fig-maps
#| fig-cap: "Accuracy (top) and precision (bottom) per station, reference set, ERA5 single-levels. Colours clipped at ±30 m and 15 m."
fig("A_maps.png")
```

```{r}
#| label: fig-class
#| fig-cap: "Error by station class (reference set). Terrain = sub-grid orography SD in the ERA5 cell."
fig("A_by_class.png")
```

- **The bias follows the terrain the grid can't resolve** (@fig-drivers). It grows with the gap
  between station and ERA5 orography, and the SD grows with sub-grid roughness above ~100 m.
- **Almost no cycle in time** (@fig-cycles). Hour of day, season, boundary layer height, surface
  inversions and pressure tendency each explain < 0.5% of deviance (Appendix B).
- **ERA5-Land is ~4× worse** (@fig-land): median |bias| `r h("A_cons_land_abs_bias_median")` m, MAE
  `r h("A_cons_land_mae")` m.
- **ERA5 has improved.** For the same stations, the median |bias| fell from
  `r if (!is.null(era)) r1(era[year == min(year), abs_bias_median])` m in
  `r if (!is.null(era)) min(era$year)` to
  `r if (!is.null(era)) r1(era[year == max(year), abs_bias_median])` m in
  `r if (!is.null(era)) max(era$year)`, and precision improved by about a third (@fig-era).

```{r}
#| label: fig-drivers
#| fig-cap: "(a) Bias against the gap between station elevation and ERA5 orography; dashed: the plausibility bound (30 m + 10% of the gap). (b) Precision against sub-grid terrain roughness (reference set, GAM fit)."
fig("A_drivers.png")
```

```{r}
#| label: fig-cycles
#| fig-cap: "Diurnal (3 h bins) and seasonal cycle of the de-biased error (RMS), by climate zone."
fig("A_cycles.png")
```

```{r}
#| label: fig-land
#| fig-cap: "Share of stations below a given |bias|, ERA5 single-levels vs ERA5-Land (reference set)."
#| out-width: 65%
fig("A_sl_vs_land.png")
```

```{r}
#| label: fig-era
#| fig-cap: "Change over time, for the stations reporting in every year."
#| out-width: 65%
fig("A_era.png")
```

## In flight

```{r}
#| label: tbl-flight
#| tbl-cap: "In-flight error of GeoPressureR, relative to the surface level of the same sounding (m). Last row: averaged over the heights at which tracked birds fly."
sel <- c("(500,1e+03]" = "0.5–1 km", "(1e+03,1.5e+03]" = "1–1.5 km", "(2e+03,3e+03]" = "2–3 km",
  "(3e+03,4e+03]" = "3–4 km", "(5e+03,6e+03]" = "5–6 km")
fl <- Bsl[hbin %in% names(sel), .(`height above ground` = sel[hbin], bias = r1(bias),
  SD = r1(sd), MAE = r1(mae), RMSE = r1(rmse))]
bw <- tab("B_bird_weighted.csv")[method == "GeoPressureR (ERA5 single-levels)"]
fl <- rbind(fl, data.table(`height above ground` = "**bird-weighted**", bias = r1(bw$bias),
  SD = "", MAE = r1(bw$mae), RMSE = r1(bw$rmse)))
knitr::kable(fl)
```

- **Altitude is too low, by ~1–1.5% of height** (@fig-height). The bias grows steadily with height,
  and the scatter grows faster.
- **The temperature profile explains it** (@fig-temperature). The error predicted from the sonde's
  own temperature profile explains `r round(100 * te$r2)`% of the variance
  (r = `r round(te$r, 2)`). The formula's layer is usually too cold (shallower real lapse rates,
  inversions, humidity), so the bird is placed too low.
- **Worst in continental and polar winters** (@fig-climate), where strong surface inversions sit
  under much warmer air.
- **ERA5-Land behaves the same in flight.** Its orography offset cancels relative to the surface,
  which confirms that its problem is a fixed spatial offset.

```{r}
#| label: fig-height
#| fig-cap: "Bias (a) and SD (b) of the in-flight error against height above ground, with the height distribution of geolocator flight points (right)."
fig("B_height.png")
```

```{r}
#| label: fig-temperature
#| fig-cap: "Observed in-flight error against the error predicted from the difference between the assumed and the observed layer temperature (levels above 200 m). Red: 1:1 line."
#| out-width: 65%
fig("B_temperature.png")
```

```{r}
#| label: fig-climate
#| fig-cap: "In-flight bias by climate zone and season (solid: summer half-year; dashed: winter half-year; southern hemisphere shifted by six months)."
#| out-width: 80%
fig("B_height_climate.png")
```

## Improving the formula

```{r}
#| label: tbl-formula
#| tbl-cap: "Bird-weighted in-flight error on held-out stations (mean ± SD over 20 random 50/50 station splits; the lapse rate is fitted on the other half). MAE gain: MAE reduction vs the current formula on the same stations, mean (worst split), m."
knitr::kable(fcv[, .(formula = method,
  `L (K/km)` = sprintf("%.2f ± %.2f", 1000 * L_mean, 1000 * L_sd),
  bias = sprintf("%.1f ± %.1f", bias_mean, bias_sd), MAE = sprintf("%.1f ± %.1f", mae_mean, mae_sd),
  RMSE = sprintf("%.1f ± %.1f", rmse_mean, rmse_sd),
  `MAE gain` = sprintf("%.1f (%.1f)", mae_gain_mean, mae_gain_min))])
```

- **Fitted lapse rate: `r sprintf("%.1f", L_t2m)` K/km with dry air, `r sprintf("%.1f", L_tv)` K/km
  with virtual temperature.** It is well determined: ±`r sprintf("%.1f", ff[temperature == "Tv", L_cv_sd])` K/km
  across splits, and `r sprintf("%.1f", ff[temperature == "Tv", L_equal_bins])` K/km when all height
  bins weigh equally. About a third of the gap to −6.5 K/km is humidity.
- **Virtual temperature alone** reduces the held-out bias from `r fc("current", "bias_mean")` m to
  `r fc("tv", "bias_mean")` m. The correction ranges from `r sprintf("%.1f", min(ftv$tv_minus_t))` K
  (polar winter) to `r sprintf("%.1f", max(ftv$tv_minus_t))` K (tropics).
- **Both together**: bias `r fc("tv_lapse", "bias_mean")` m, MAE `r fc("current", "mae_mean")` →
  `r fc("tv_lapse", "mae_mean")` m, better in all 20 splits. Held-out and in-sample errors are
  practically identical.
- **Fitting the lapse rate alone performs about as well** statistically (@tbl-formula).
- **The underestimation disappears up to ~3 km** (@fig-formula):
  `r fhb("tv_lapse", "(1e+03,1.5e+03]")` m at 1–1.5 km and `r fhb("tv_lapse", "(2e+03,3e+03]")` m
  at 2–3 km, vs `r fhb("current", "(1e+03,1.5e+03]")` and `r fhb("current", "(2e+03,3e+03]")` m.
  Above that, the corrected formula slightly overestimates (`r fhb("tv_lapse", "(5e+03,6e+03]")` m
  at 5–6 km).
- **The scatter does not change.** A constant lapse rate can't follow the day-to-day profile.
- **Regional and seasonal biases remain** (@fig-formula-climate). Fitted per group, the lapse rate
  ranges from `r sprintf("%.1f", min(fgrp$L_tv))` K/km (arid summer) to
  `r sprintf("%.1f", max(fgrp$L_tv))` K/km (polar winter). With the global value, continental and
  polar winters stay too low and arid summers become too high.
- **No change on the ground.** Pooled MAE goes from `r sprintf("%.2f", fg$mae)` to
  `r sprintf("%.2f", fg$mae_new)` m; the largest change of any single observation is
  `r sprintf("%.1f", fg$max_abs_change)` m.

```{r}
#| label: fig-formula
#| fig-cap: "Bias (a) and SD (b) against height for the current formula and the three variants (lapse rates fitted on all stations), with the height distribution of geolocator flight points."
fig("B_formula.png")
```

```{r}
#| label: fig-formula-climate
#| fig-cap: "In-flight bias by climate zone and season: current formula (left), and virtual temperature with the fitted lapse rate (right). Solid: summer half-year; dashed: winter half-year."
fig("B_formula_climate.png")
```

# Discussion

## What users can expect

| Situation | Typical error | Driver |
|---|---|---|
| Absolute altitude on the ground, flat terrain | bias `r t("flat (<20 m)", "abs_bias_median")` m (p90 `r t("flat (<20 m)", "abs_bias_p90")` m), MAE `r t("flat (<20 m)", "mae")` m | ERA5 surface pressure at 0.25° |
| Absolute altitude on the ground, mountains | bias `r t("mountain (>300 m)", "abs_bias_median")` m (p90 `r t("mountain (>300 m)", "abs_bias_p90")` m), MAE `r t("mountain (>300 m)", "mae")` m | terrain the 0.25° grid can't resolve |
| Altitude *changes* at one site | SD `r t("flat (<20 m)", "sd_median")` m (flat) to `r t("mountain (>300 m)", "sd_median")` m (mountains) | as above; almost no diurnal or seasonal cycle |
| In flight, 1–1.5 km above ground | `r h("B_rel_bias_1000_1500", 0)` m bias, RMSE `r h("B_rel_rmse_1000_1500", 0)` m (on top of the ground error) | assumed temperature profile |
| In flight, 2–3 km above ground | `r h("B_rel_bias_2000_3000", 0)` m bias, RMSE `r h("B_rel_rmse_2000_3000", 0)` m | as above; ~2× in continental/polar winter |

## Key points

- **On the ground, the error is a fixed offset per place, not noise.** Station identity explains
  almost all the explainable variation. Relative altitudes at one site (SD ~4 m) are therefore more
  reliable than absolute ones (MAE ~7 m).
- **In flight, the error is systematic and grows with height.** Flight altitudes from GeoPressureR
  are conservative. A bird at 2 km above ground is placed about 25–35 m too low, with a scatter of
  30–50 m; in a continental or polar winter, both roughly double.
- **Older tracks are somewhat less accurate.** ERA5's ground-level bias in 1990 was about twice that
  of 2024.
- **Use ERA5 single-levels, not ERA5-Land, for altitude.** ERA5-Land's orography is inconsistent
  with its surface pressure, which shifts every altitude by a fixed, place-dependent offset
  ([GeoPressureAPI#27](https://github.com/GeoPressure/GeoPressureAPI/issues/27)).

## Recommendation for GeoPressureR

- **Add the virtual temperature.** It corrects a simplification, has no free parameter, and on its
  own removes about 40% of the in-flight bias. It only needs the ERA5 2 m dewpoint.
- **Combine it with a lapse rate of `r sprintf("%.1f", L_tv)` K/km.** The held-out in-flight bias
  drops from `r fc("current", "bias_mean")` m to `r fc("tv_lapse", "bias_mean")` m, with no effect on
  the ground. The estimate is stable across splits and objectives, and it lies within the range
  reported near the surface.
- **Prefer the combination to a lapse rate alone.** A dry-air fit (`r sprintf("%.1f", L_t2m)` K/km)
  does about as well statistically. But it folds humidity and the temperature profile into one
  constant with no physical meaning, which is harder to interpret and to refine.
- **Beyond a constant lapse rate.** The remaining scatter and regional biases come from the actual
  temperature profile (inversions, deep convective boundary layers). Fixing them needs a lapse rate
  that varies with place and season, or the ERA5 temperature profile or geopotential on pressure
  levels. The latter could remove most of the temperature-driven variance
  (`r round(100 * te$r2)`% of the in-flight error), but it needs far more ERA5 data per point.

## Limitations

- **Not independent of ERA5.** ERA5 assimilates both the station pressures and the radiosondes, so
  errors *at these sites* are optimistic. Remote areas are likely somewhat worse, especially for
  ground-level bias.
- **The sensor is not included.** These are errors of the retrieval, given a perfect pressure reading.
- **Radiosonde geometry.** The ERA5 column at the launch site is used, and balloon drift (a few km
  below 5 km) is ignored. Sonde heights are themselves hypsometric. The ~10 m scatter in the first
  100 m partly reflects integer-metre reporting.
- **Reference screen.** `r hi("A_n_excluded")` stations were excluded. Neighbours and DEMs support
  this where available, but not every excluded station could be checked. The reference set still
  holds small undetected reference errors.
- **Sampling.** Coverage is still sparse in parts of Africa, South America and over the oceans (Tier A
  is land only).

# Data and code availability

- **Code:** <https://github.com/GeoPressure/altitude-validation>.
- **ERA5 and ERA5-Land:** Copernicus Climate Change Service, via the ECMWF ARCO archive.
- **HadISD:** Met Office Hadley Centre, Non-Commercial Government Licence. It may contain data
  governed by WMO Resolution 40 Annex 1; only derived statistics are published here.
- **IGRA2:** NOAA NCEI.
- **Bird heights:** GeoLocator master data package (<https://doi.org/10.5281/zenodo.18187092>).
- **DEMs:** SRTM GL1 and ASTER GDEM v3, via OpenTopoData.

# References

- Bolton D (1980) The computation of equivalent potential temperature. *Monthly Weather Review*
  108:1046–1053. <https://doi.org/10.1175/1520-0493(1980)108%3C1046:TCOEPT%3E2.0.CO;2>
- Dunn RJH, Willett KM, Thorne PW, et al. (2012) HadISD: a quality-controlled global synoptic report
  database for selected variables at long-term stations from 1973–2011. *Climate of the Past*
  8:1649–1679. <https://doi.org/10.5194/cp-8-1649-2012>
- Dunn RJH, Willett KM, Parker DE, Mitchell L (2016) Expanding HadISD: quality-controlled, sub-daily
  station data from 1931. *Geoscientific Instrumentation, Methods and Data Systems* 5:473–491.
  <https://doi.org/10.5194/gi-5-473-2016>
- Durre I, Vose RS, Wuertz DB (2006) Overview of the Integrated Global Radiosonde Archive. *Journal
  of Climate* 19:53–68. <https://doi.org/10.1175/JCLI3594.1>
- Durre I, Yin X, Vose RS, et al. (2018) Enhancing the data coverage in the Integrated Global
  Radiosonde Archive. *Journal of Atmospheric and Oceanic Technology* 35:1753–1770.
  <https://doi.org/10.1175/JTECH-D-17-0223.1>
- ECMWF (2025) Guidance on SYNOP surface pressure monitoring issues. ECMWF Confluence (E. Kuscu).
  <https://confluence.ecmwf.int/pages/viewpage.action?pageId=298952900>
- Hersbach H, Bell B, Berrisford P, et al. (2020) The ERA5 global reanalysis. *Quarterly Journal of
  the Royal Meteorological Society* 146:1999–2049. <https://doi.org/10.1002/qj.3803>
- ICAO (1993) *Manual of the ICAO Standard Atmosphere*, Doc 7488/3, 3rd ed. International Civil
  Aviation Organization, Montreal.
- ISO (1975) *ISO 2533:1975 Standard Atmosphere*. International Organization for Standardization.
- Minder JR, Mote PW, Lundquist JD (2010) Surface temperature lapse rates over complex terrain:
  lessons from the Cascade Mountains. *Journal of Geophysical Research* 115:D14122.
  <https://doi.org/10.1029/2009JD013493>
- Muñoz-Sabater J, Dutra E, Agustí-Panareda A, et al. (2021) ERA5-Land: a state-of-the-art global
  reanalysis dataset for land applications. *Earth System Science Data* 13:4349–4383.
  <https://doi.org/10.5194/essd-13-4349-2021>
- NASA/METI/AIST/Japan Spacesystems and U.S./Japan ASTER Science Team (2019) ASTER Global Digital
  Elevation Model V003. NASA EOSDIS Land Processes DAAC. <https://doi.org/10.5067/ASTER/ASTGTM.003>
- NASA JPL (2013) NASA Shuttle Radar Topography Mission Global 1 arc second. NASA EOSDIS Land
  Processes DAAC. <https://doi.org/10.5067/MEaSUREs/SRTM/SRTMGL1.003>
- Rodríguez E, Morris CS, Belz JE (2006) A global assessment of the SRTM performance.
  *Photogrammetric Engineering & Remote Sensing* 72:249–260. <https://doi.org/10.14358/PERS.72.3.249>
- Rolland C (2003) Spatial and seasonal variations of air temperature lapse rates in Alpine regions.
  *Journal of Climate* 16:1032–1046.
  <https://doi.org/10.1175/1520-0442(2003)016%3C1032:SASVOA%3E2.0.CO;2>
- Smith A, Lott N, Vose R (2011) The Integrated Surface Database: recent developments and
  partnerships. *Bulletin of the American Meteorological Society* 92:704–708.
  <https://doi.org/10.1175/2011BAMS3015.1>
- Stone PH, Carlson JH (1979) Atmospheric lapse rate regimes and their parameterization. *Journal
  of the Atmospheric Sciences* 36:415–423.
  <https://doi.org/10.1175/1520-0469(1979)036%3C0415:ALRRAT%3E2.0.CO;2>
- Vasiljevic D, Andersson E, Isaksen L, Garcia-Mendez A (2006) Surface pressure bias correction in
  data assimilation. *ECMWF Newsletter* 108:20–27. <https://doi.org/10.21957/uv295rfmx5>
- Wallace JM, Hobbs PV (2006) *Atmospheric Science: An Introductory Survey*, 2nd ed., ch. 3.
  Academic Press. <https://doi.org/10.1016/B978-0-12-732951-2.50008-9>
- WMO (1968) *Methods in Use for the Reduction of Atmospheric Pressure*. Technical Note 91,
  WMO-No. 226. World Meteorological Organization, Geneva.
- WMO (2023) *Guide to Instruments and Methods of Observation*, Volume I (WMO-No. 8), chapters 1
  (station elevation) and 3 (atmospheric pressure). World Meteorological Organization, Geneva.

# Appendix A: reference screen

```{r}
dem_ok <- sa[is.finite(dem_srtm_centre) & is.finite(dem)]
dem_sa <- sa[is.finite(dem_srtm_centre) & is.finite(dem_aster_centre)]
rr <- sa[reference == TRUE & dem_refutes == TRUE]
rr[, dem_best := fifelse(is.finite(dem_srtm_centre), dem_srtm_centre, dem_aster_centre)]
```

## Why the DEM is not used as a filter

- **Coordinates are coarse.** `r pct(mean(sa$minute_coord, na.rm = TRUE))` fall on whole
  arc-minutes (±0.5′, up to ±900 m). A single DEM point can then be off by tens to hundreds of
  metres in rough terrain.
- **The DEM is therefore sampled on a 5×5 grid over that uncertainty**, in SRTM and ASTER GDEM. An
  elevation is "refuted" if it lies more than 15 m outside the DEM range in every available DEM.
- **SRTM would not have helped.** The Mapzen tiles used before are SRTM wherever SRTM exists (median
  difference `r round(median(abs(dem_ok$dem_srtm_centre - dem_ok$dem)))` m).
- **DEMs are only good to ~15 m.** SRTM's 90% absolute height error is 5–9 m (Rodríguez et al.
  2006). SRTM and ASTER differ by a median
  `r round(median(abs(dem_sa$dem_srtm_centre - dem_sa$dem_aster_centre)))` m (p90
  `r round(quantile(abs(dem_sa$dem_srtm_centre - dem_sa$dem_aster_centre), 0.9))` m).
- **A DEM filter would remove good stations.** At `r nrow(rr)` reference stations the DEM
  contradicts the listed elevation (median difference
  `r round(median(abs(rr$elev - rr$dem_best), na.rm = TRUE))` m), yet ERA5 reproduces it (median
  |bias| `r r1(median(abs(rr$bias_sl)))` m). There, the coordinates are wrong, not the elevation.

## Sensitivity to the screen

```{r}
#| label: tbl-sensitivity
#| tbl-cap: "Ground-level statistics (m) for looser and stricter plausibility bounds (|bias| ≤ a + b × gap), with the step-change stations kept or excluded. Used in the report: '30 m + 10%' with step-change stations excluded."
knitr::kable(sens[, .(`offset rule` = rule, `step stations` = steps, stations = n_stations,
  `median |bias|` = r1(abs_bias_median), `p90 |bias|` = r1(abs_bias_p90),
  `median SD` = r1(sd_median), MAE = r1(mae), RMSE = r1(rmse), `p95 |err|` = r1(p95_abs))])
```

- **The medians don't move.** The median bias and median SD are insensitive to the screen.
- **MAE, RMSE and p95 do.** A few stations with offsets of hundreds of metres dominate them when
  included. Loosening the bound to 60 m + 20% adds back
  `r sens[rule == "60 m + 20%" & steps == "excluded", n_stations] - sens[rule == "30 m + 10% (used)" & steps == "excluded", n_stations]`
  stations and raises the MAE from
  `r r1(sens[rule == "30 m + 10% (used)" & steps == "excluded", mae])` to
  `r r1(sens[rule == "60 m + 20%" & steps == "excluded", mae])` m.

## Basis of the plausibility bound

```{r}
#| label: tbl-basis-flat
#| tbl-cap: "|bias| where ERA5 needs no extrapolation: sub-grid SD < 20 m and station within 30 m of ERA5 orography (m)."
knitr::kable(bf[, lapply(.SD, function(x) round(x, 1))])
```

```{r}
#| label: tbl-basis-gap
#| tbl-cap: "Extrapolation error over a height gap, measured by radiosondes: per-station mean error relative to the surface level, divided by height above ground."
knitr::kable(bg[, .(`height (m)` = hbin, stations = n_stations,
  `median (%)` = round(median_pct, 2), `p99 of |.| (%)` = round(p99_abs_pct, 1),
  `max |.| (%)` = round(max_abs_pct, 1))])
```

## Excluded stations

```{r}
#| label: tbl-excluded
#| tbl-cap: "Stations excluded from the reference set (m). SRTM/ASTER: DEM at the listed coordinates. DEM refutes: listed elevation more than 15 m outside the DEM range of the coordinate-uncertainty box in every available DEM. Neighbour: median ERA5 bias in 2023 at up to two HadISD stations within 50 km (blank: none, or not tested)."
ex[!grepl("step", reason), `:=`(step = NA, step_at = NA)]
knitr::kable(ex[, .(station = name, reason, `elev.` = elev, `ERA5 orog.` = z_sl, bias, SD = sd,
  step, `from` = step_at, SRTM = dem_srtm, ASTER = dem_aster, `DEM refutes` = dem_refutes,
  `neighbour bias` = nb_bias_median, `neighbour km` = nb_km_min)])
```

# Appendix B: supplementary tables

```{r}
#| label: tbl-class
#| tbl-cap: "Ground-level error by station class (reference set, ERA5 single-levels, m)."
knitr::kable(A[group != "all" & dataset == "sl", .(group, class = subset, stations = n_stations,
  `median |bias|` = r1(abs_bias_median), `p90 |bias|` = r1(abs_bias_p90),
  `median SD` = r1(sd_median), `p90 SD` = r1(sd_p90), MAE = r1(mae), `p95 |err|` = r1(p95_abs))])
```

```{r}
#| label: tbl-gam
#| tbl-cap: "Drivers: deviance explained by the full GAM, and lost when each term is dropped. Station level: log |bias| and log SD. Observation level: de-biased absolute error (GAM with a station random effect)."
d1 <- tab("A_drivers_station.csv")
d2 <- tab("A_drivers_obs.csv")[, response := "|de-biased error| (observations)"]
d <- rbind(d1, d2, fill = TRUE)
knitr::kable(d[, .(response, term, `full model` = round(dev_expl_full, 3),
  `lost when dropped` = round(dev_expl_drop, 4), n = format(n, big.mark = ","))])
```

```{r}
#| label: tbl-height
#| tbl-cap: "In-flight error by height, relative to the surface level of the same sounding (m)."
knitr::kable(B[!is.na(hbin) & hbin != "", .(dataset = fifelse(method == "ERA5-Land", "ERA5-Land",
  "single-levels"), `height (m)` = hbin, levels = format(n, big.mark = ","), stations = n_stations,
  bias = r1(bias), SD = r1(sd), MAE = r1(mae), RMSE = r1(rmse), `p95 |err|` = r1(p95_abs))])
```

```{r}
#| label: tbl-daynight
#| tbl-cap: "In-flight bias by height, day/night and half-year (GeoPressureR, m)."
d <- tab("B_height_daynight_season.csv")[method == "GeoPressureR (ERA5 single-levels)"]
knitr::kable(dcast(d, hbin ~ daynight + season, value.var = "bias",
  fun.aggregate = function(x) round(x, 1)))
```

```{r}
#| label: tbl-formula-height
#| tbl-cap: "In-flight bias by height for the formula variants (m)."
knitr::kable(
  dcast(fh[, .(code = factor(code, c("current", "tv", "lapse", "tv_lapse")), agl_mid, hbin,
    bias = round(bias, 1))], agl_mid + hbin ~ code, value.var = "bias")[, !"agl_mid"],
  col.names = c("height (m)", fbw$method[match(c("current", "tv", "lapse", "tv_lapse"), fbw$code)])
)
```

```{r}
#| label: tbl-lapse-group
#| tbl-cap: "Lapse rate fitted separately for each climate zone and half-year."
knitr::kable(fgrp[, .(climate, season, stations = n_st, `L with T2m (K/km)` = round(L_t2m, 1),
  `L with Tv (K/km)` = round(L_tv, 1))])
```
