---
title: "How accurate is altitude from pressure geolocators?"
subtitle: "A global validation of GeoPressureR altitude against station barometers and radiosondes"
author: "Raphaël Nussbaumer"
date: today
format:
  gfm:
    toc: false
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 9
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: ".."
---

```{r setup}
invisible(Sys.setlocale("LC_CTYPE", "en_US.UTF-8"))
library(data.table)
tab <- function(f) fread(file.path("output/tables", f))
hl <- tab("headline.csv")
h <- function(k, digits = 1) {
  v <- hl[metric == k, value]
  if (!length(v) || is.na(v)) return("NA")
  formatC(v, format = "f", digits = digits, big.mark = ",")
}
hi <- function(k) h(k, 0)
# Paths relative to report/, so the figures resolve both on GitHub (index.md) and in the HTML.
fig <- function(f) knitr::include_graphics(file.path("..", "output/figures", f), error = FALSE)
A <- tab("A_summary.csv")
B <- tab("B_height.csv")
gp_version <- readLines("output/tables/geopressurer_version.txt")
# Formula variants (07b_formula.R)
ff <- tab("formula_fit.csv")
fcv <- tab("formula_cv.csv")
fbw <- tab("formula_bird_weighted.csv")
fh <- tab("formula_height.csv")
fg <- tab("formula_ground.csv")
fgrp <- tab("formula_lapse_by_group.csv")
ftv <- tab("formula_tv_correction.csv")
L_tv <- ff[temperature == "Tv", L_fit]
L_t2m <- ff[temperature == "T2m", L_fit]
fb <- function(k, col, digits = 1) formatC(fbw[code == k][[col]], format = "f", digits = digits)
fc <- function(k, col, digits = 1) formatC(fcv[code == k][[col]], format = "f", digits = digits)
fhb <- function(k, bin) round(fh[code == k & hbin == bin, bias])
```

::: {.callout-note appearance="simple"}
## In short

**At ground level**, GeoPressureR's default (ERA5 single-levels) retrieves a geolocator's altitude
with a median per-site **bias of `r h("A_cons_abs_bias_median")` m** (90% of sites within
`r h("A_cons_abs_bias_p90")` m) and a **temporal scatter (SD) of `r h("A_cons_sd_median")` m**
(`r hi("A_cons_n_stations")` stations worldwide, 2022–2024). Pooled over all station-hours the mean
absolute error is **`r h("A_cons_mae")` m** and 95% of errors are below `r h("A_cons_p95")` m.
Terrain roughness is the main driver: errors roughly triple from flat to mountainous terrain.

**In flight**, altitude is **underestimated by about 1–1.5% of the height above the ground**
(`r h("B_rel_bias_1000_1500", 0)` m at 1–1.5 km, `r h("B_rel_bias_2000_3000", 0)` m at 2–3 km,
`r h("B_rel_bias_5000_6000", 0)` m at 5–6 km), because the formula assumes a standard temperature
profile. The error is largest in continental and polar winters (surface inversions). Averaged over
the heights at which tracked birds fly, flight adds a mean absolute error of
**`r h("B_bird_weighted_mae")` m**.

**ERA5-Land** is much worse for absolute altitude (MAE `r h("A_cons_land_mae")` m) and should not be
used for it.

**The formula can be improved.** Using the virtual temperature (which accounts for humidity) and a
lapse rate of `r sprintf("%.1f", L_tv)` K/km estimated from the radiosondes reduces the in-flight
bias from `r fb("current", "bias")` m to `r fb("tv_lapse", "bias")` m on stations not used for the
fit, without changing ground-level altitude. It is not (yet) what GeoPressureR does.
:::

# Why this study

GeoPressureR (and GeoPressureAPI behind it) converts the pressure a geolocator records into an
altitude using ERA5 reanalysis: the surface pressure $p_0$ and 2 m temperature $T_0$ at the bird's
location and hour, the ERA5 orography $z_0$, and the barometric formula with a standard lapse rate
$L = -6.5$ K/km,

$$
z = z_0 + \frac{T_0}{L}\left[\left(\frac{p}{p_0}\right)^{-R L / (g M)} - 1\right].
$$

Users want to know how good that altitude is, where and when it is worse, and why. An earlier check
over the Alps
([GeoPressureAPI#27](https://github.com/GeoPressure/GeoPressureAPI/issues/27)) found ~9 m mean
absolute error with ERA5 single-levels and led to making single-levels the default. This study extends
that check to the whole globe, a full annual cycle, three recent years, and — using radiosondes —
to the heights at which birds fly.

# Methods

## What is validated

The exact GeoPressureR code path is evaluated: grid-cell snapping, nearest-hour matching, the ERA5
orography files and `pressure_to_altitude()` are taken from GeoPressureR `r gp_version`, reading
ERA5 from the ECMWF ARCO archive (`pressurepath_create(source = "arco")`). A cross-check against
GeoPressureAPI (Earth Engine) on `r if (file.exists("output/tables/api_crosscheck_summary.csv")) tab("api_crosscheck_summary.csv")$n else "NA"`
station-hours found a maximum difference of `r h("api_max_abs_diff", 2)` m, so every result below
applies to both entry points.

Two ERA5 products are compared: **ERA5 single-levels** (0.25°, the default) and **ERA5-Land**
(0.1°). The mixed `"both"` option is not evaluated separately: over land it is ERA5-Land.

## Altitude formula and alternatives

**Why a formula error grows with height.** In hydrostatic balance, the thickness of the layer
between the surface pressure $p_0$ and the pressure $p$ is given by the hypsometric equation
(Wallace & Hobbs 2006, ch. 3)

$$
z - z_0 = \frac{R_d}{g}\,\bar T_v \,\ln\frac{p_0}{p},
$$

where $R_d = R/M$ is the gas constant of dry air and $\bar T_v$ the (log-pressure weighted) mean
*virtual* temperature of the layer. GeoPressureR's formula is this equation for a particular
assumed profile: a temperature that starts at the ERA5 2 m temperature $T_0$ and decreases linearly
at the lapse rate $L = -6.5$ K/km of the ICAO standard atmosphere (ICAO 1993; ISO 2533:1975), with
dry air ($T_v = T$). Because the thickness is proportional to $\bar T_v$, an error in the assumed
mean temperature translates into the same *relative* error in height above the ground,

$$
\frac{\delta (z - z_0)}{z - z_0} = \frac{\bar T_\text{assumed} - \bar T_{v,\text{true}}}{\bar T_{v,\text{true}}},
$$

so 3 K (≈1%) too cold puts a bird 2 km above the ground about 20 m too low. On the ground,
$z - z_0$ is only the gap between the station and the ERA5 orography (median
`r round(fg$median_abs_dz)` m), so the temperature assumption hardly matters there.

The assumed layer is too cold for two separate reasons, each addressed by one change:

1. **Virtual temperature.** Water vapour is lighter than dry air, so moist air occupies a thicker
   layer than dry air at the same temperature. The hypsometric equation requires the virtual
   temperature $T_v = T\,(1 + 0.608\,q)$, where $q$ is the specific humidity; using $T$ is a
   simplification, not a modelling choice. We compute $q$ from the ERA5 2 m dewpoint and surface
   pressure, with the saturation vapour pressure of Bolton (1980), and replace $T_0$ by $T_{v,0}$.
   The correction has no free parameter. At the radiosonde stations it averages
   `r sprintf("%.1f", min(ftv$tv_minus_t))` K in polar winter and up to
   `r sprintf("%.1f", max(ftv$tv_minus_t))` K in the tropics (0.1–1.2% of height). As in pressure reduction to sea level (WMO 1968), humidity at
   the surface is assumed to hold through the layer, which slightly overestimates it aloft.
2. **Lapse rate.** −6.5 K/km is a mean over the whole troposphere (0–11 km). Lapse rates in the
   lowest kilometres vary with latitude, season and time of day (Stone & Carlson 1979; Rolland
   2003) and are often shallower: for example 3.9–5.2 K/km annual mean near-surface lapse rates in
   the Cascade Mountains (Minder et al. 2010), and much shallower still, or inverted, in winter
   inversions. Rather than adopt a value from the literature, we **estimate $L$ from the
   radiosondes**.

**Estimating and testing the lapse rate.** $L$ is the value that minimises the bird-weighted mean
squared error relative to the surface level (Tier B; height bins weighted by the share of
geolocator flight points, stations weighted equally), found by one-dimensional optimisation. It is
estimated separately with $T_0$ and with $T_{v,0}$. Its robustness is tested in three ways:
(i) 20 repeated random 50/50 splits of the stations, fitting on one half and evaluating on the
other; (ii) separate fits for each climate zone and half-year; (iii) an alternative objective that
weights all height bins equally. Its effect at ground level is evaluated on Tier A. These variants
are evaluated here only; all other results use the formula as implemented in GeoPressureR.

## Reference data

**Tier A — surface barometers (ground level).** HadISD v3.4.3.2025f (Dunn et al. 2012, 2016), a
quality-controlled subset of NOAA ISD. Its station-level pressure is treated as a geolocator reading
and the retrieved altitude is compared with the station elevation. Stations were chosen for spatial
balance: in every 5°×5° cell the highest, the lowest and one random station, plus every station
above 1000 m. Stations had to report station pressure on at least half the days of each of
2022–2024, at least 6-hourly; the set was then thinned to one random station per 5° cell plus one
per 2.5° cell above 1000 m. ERA5-Land was evaluated on a random 200 of these stations, and a random
100 reporting in 1990, 2005 and 2015 test for changes over time.

**Tier B — radiosondes (0–6 km above ground).** IGRA2 (Durre et al. 2006, 2018), 2023–2024. At
every level the sonde reports pressure and geopotential height; the pressure is given to the
altitude formula at the launch site and hour of that level, and the retrieved altitude is compared
with the sonde's height. Stations: the highest and one random active station per 10°×10° cell, plus
all above 1000 m.

## Separating accuracy from precision, and handling errors in the reference

At ~5–10 m, errors in the reference itself matter. Three choices keep them out of the headline
numbers:

- **Accuracy vs precision.** Each station's error is split into its mean (the *bias*: constant in
  time, relevant for absolute altitude) and the SD around that mean (*precision*: relevant for
  altitude changes). A wrong station elevation shifts only the bias, so precision is robust to
  metadata errors.
- **Reference screen (Tier A).** Stations whose own elevation or pressure cannot be trusted are
  excluded from the reference set (below).
- **Height relative to the surface level (Tier B).** The error at a level minus the error at the same
  sounding's surface level cancels the station elevation entirely, isolating how the error grows
  with height.

Gross observation errors were flagged when more than 100 m (Tier A) or 150 m (Tier B) and 10 robust
SD from the station's (or height bin's) median error: `r h("A_gross_fraction", 5)` of Tier A and
`r h("B_gross_fraction", 5)` of Tier B observations.

Pooled statistics give every station equal weight.

### Screening the surface stations

```{r screen}
sa <- tab("A_station_stats.csv")
ex <- tab("A_excluded.csv")
nbp <- tab("A_neighbours.csv")
bf <- tab("A_screen_basis_flat.csv")
bg <- tab("A_screen_basis_gap.csv")
ex_nb <- ex[is.finite(nb_bias_median)]
ref_nb <- nbp[id %in% sa[reference == TRUE, id] & abs(bias) > 15,
  .(bias = bias[1], nb = median(nb_bias)), by = id]
n_shared <- ref_nb[sign(nb) == sign(bias) & abs(nb) >= 0.5 * abs(bias), .N]
imp <- sa[consistent == FALSE]
dem_ok <- sa[is.finite(dem_srtm_centre) & is.finite(dem)]
dem_sa <- sa[is.finite(dem_srtm_centre) & is.finite(dem_aster_centre)]
rr <- sa[reference == TRUE & dem_refutes == TRUE]
rr[, dem_best := fifelse(is.finite(dem_srtm_centre), dem_srtm_centre, dem_aster_centre)]
pct <- function(x) paste0(round(100 * x), "%")
nm <- function(x) paste(tools::toTitleCase(tolower(x)), collapse = ", ")
```

A station's listed elevation and reported station pressure are not always what they seem, and this
is well documented. At ECMWF, surface pressure reports from several hundred SYNOP stations are
biased, often by several hPa. These biases are "mostly related to incorrect assumptions about the
station heights" and stay fairly constant in time (Vasiljevic et al. 2006). ECMWF's monitoring
guidance lists the usual causes (ECMWF 2025):

- wrong latitude, longitude or elevation metadata;
- GNSS-derived elevations without the geoid correction;
- stations that report only mean sea-level pressure;
- sensor or encoding errors;
- step changes when the barometer height or the station position changes.

WMO distinguishes the datum to which station pressure refers ($H_p$) from the ground elevation of
the station, and at aerodromes from the aerodrome elevation (WMO 2023). A listed elevation can
therefore legitimately differ from the height the pressure refers to. HadISD takes its station
metadata from ISD (Smith et al. 2011). It is quality-controlled but not homogenised, and it merges
records of co-located stations (Dunn et al. 2012, 2016). ERA5 bias-corrects or rejects such
stations when it assimilates them (Vasiljevic et al. 2006; Hersbach et al. 2020), so where a single
station and ERA5 disagree by a constant offset, the station is usually at fault.

Two rules remove such stations from the **reference set** used for all statistics except the "all
stations" row:

1. **Implausible offset** (`r hi("A_n_inconsistent")` stations): |bias| > 30 m + 10% of the gap
   between station elevation and ERA5 orography. The bound is set from data that do not use the
   bias of the station being tested (Appendix):
   - Where ERA5 needs no extrapolation (flat terrain, gap < 30 m, `r bf$n` stations), 95% of
     stations are within `r round(bf$p95)` m, so 30 m is about twice that.
   - Over a height gap, the radiosondes measure the extrapolation error directly. The per-station
     mean is `r round(max(bg$median_pct), 1)`% to `r round(min(bg$median_pct), 1)`% of the gap, and
     99% of stations stay within `r round(min(bg$p99_abs_pct))`–`r round(max(bg$p99_abs_pct))`%.
     The 10% allowance is therefore generous.
2. **Step change** (`r hi("A_n_step")` stations): a clear jump in the station's error during
   2022–2024. Monthly median errors are fitted with a month-of-year effect plus one step at the
   best breakpoint; a station is flagged if |step| ≥ 10 m and |t| ≥ 10. ERA5 does not jump at one
   site, but a station does when it moves, changes barometer or changes its reference height (ECMWF
   2025). Examples are `r nm(head(ex[reason == "step change"][order(-abs(step))]$name, 3))`
   (`r paste(head(ex[reason == "step change"][order(-abs(step))]$step, 3), collapse = ", ")` m).

In total `r hi("A_n_excluded")` of `r hi("A_n_stations")` stations
(`r pct(as.numeric(h("A_n_excluded", 0)) / as.numeric(h("A_n_stations", 0)))`) are excluded.
The Appendix lists them.

**Independent checks that the screen removes reference errors, not ERA5 errors.**

- *Neighbours.* ERA5's bias was computed in 2023 at up to two other HadISD stations within 50 km of
  every station with |bias| > 15 m. `r nrow(ex_nb)` excluded stations have such a neighbour. ERA5
  matches the neighbours of all of them to within `r ceiling(max(abs(ex_nb$nb_bias_median)))` m,
  while the excluded stations themselves are off by up to `r format(max(abs(ex_nb$bias)), big.mark = ",")` m.
  Their offsets belong to the station, not to ERA5 (the same logic ECMWF uses to decide whether a
  station or the model is biased; Vasiljevic et al. 2006). Conversely, `r n_shared` of the
  `r nrow(ref_nb)` tested reference stations with |bias| > 15 m share their offset with a
  neighbour. These are genuine ERA5 errors (mostly in mountains), and the screen keeps them.
- *No mechanism in ERA5.* `r sum(imp$dem_refutes == FALSE, na.rm = TRUE)` of the
  `r nrow(imp)` implausible offsets are at stations whose elevation an independent DEM confirms,
  often in flat terrain close to ERA5's orography (e.g. `r nm(head(imp[dem_refutes == FALSE & abs(dz_sl) < 30][order(-abs(bias_sl))]$name, 3))`).
  There, ERA5 has nothing to extrapolate. The offset must come from the pressure's datum or from
  the barometer.
- *Constant in time.* The excluded stations track ERA5 hour by hour (median SD
  `r round(median(imp$sd_sl), 1)` m, vs `r h("A_cons_sd_median")` m in the reference set) at a fixed offset. That is the
  signature ECMWF reports for station-height errors, not what a reanalysis error would look like.

**Why the DEM is not used as a filter.** Earlier versions also reported a "DEM-trusted" subset.
The check was replaced because it measures the precision of the station coordinates more than the
elevation:

- `r pct(mean(sa$minute_coord, na.rm = TRUE))` of the coordinates fall on whole arc-minutes, i.e.
  ±0.5′ (up to ±900 m). A single DEM point can then be off by tens to hundreds of metres in rough
  terrain.
- The DEM was therefore sampled on a 5×5 grid covering that uncertainty, in SRTM and in ASTER GDEM.
  The Mapzen terrain tiles used before are SRTM wherever SRTM exists (median difference
  `r round(median(abs(dem_ok$dem_srtm_centre - dem_ok$dem)))` m), so SRTM would not have helped.
  SRTM's own 90% absolute height error is 5–9 m (Rodríguez et al. 2006), and SRTM and ASTER differ
  by a median `r round(median(abs(dem_sa$dem_srtm_centre - dem_sa$dem_aster_centre)))` m
  (p90 `r round(quantile(abs(dem_sa$dem_srtm_centre - dem_sa$dem_aster_centre), 0.9))` m). A DEM
  cannot verify an elevation to better than ~15 m.
- At `r nrow(rr)` reference stations, every available DEM contradicts the listed elevation
  (median difference `r round(median(abs(rr$elev - rr$dem_best), na.rm = TRUE))` m). Yet ERA5 reproduces the listed
  elevation (median |bias| `r round(median(abs(rr$bias_sl)), 1)` m). At those stations the
  coordinates are wrong, not the elevation, so a DEM filter would have removed good stations and
  kept bad ones.

The DEM is kept only as supporting evidence in the table of excluded stations.

**What remains.** The reference set still contains smaller reference errors that no screen can
detect: steps below 10 m, barometers a few metres above the listed ground, rounded elevations. Its
accuracy is therefore an upper bound on ERA5's own error. The Appendix shows how the
pooled statistics change with the screen. The median bias barely moves
(`r round(min(tab("A_sensitivity.csv")$abs_bias_median), 1)`–`r round(max(tab("A_sensitivity.csv")$abs_bias_median), 1)` m
across all variants), but MAE and RMSE depend on it, because a handful of stations with offsets of
hundreds of metres dominate them when included.

## Drivers

Candidate drivers of accuracy and precision: the difference between station elevation and ERA5
orography, sub-grid terrain roughness (ERA5 `sdor`), latitude, Köppen-Geiger climate zone (station
level); local solar hour, season, boundary layer height, skin − 2 m temperature (a proxy for
surface inversions) and 6 h surface pressure tendency (observation level). Their effect is
estimated with GAMs (mgcv) and ranked by the deviance explained lost when each is dropped.

For Tier B, the error expected from the temperature assumption alone follows from the relative-error
equation above: $\Delta z \cdot (\bar T_\text{assumed} - \bar T_\text{obs}) / \bar T_\text{obs}$, where
$\bar T_\text{obs}$ is the sonde's log-pressure-weighted mean (dry-bulb) temperature of the layer
below the level.

# Results: ground level (Tier A)

```{r}
knitr::kable(
  A[group == "all", .(subset, dataset = fifelse(dataset == "sl", "single-levels", "land"),
    stations = n_stations, `obs` = format(n_obs, big.mark = ","),
    `median |bias|` = round(abs_bias_median, 1), `p90 |bias|` = round(abs_bias_p90, 1),
    `median SD` = round(sd_median, 1), `p90 SD` = round(sd_p90, 1), MAE = round(mae, 1),
    RMSE = round(rmse, 1), `p95 |err|` = round(p95_abs, 1))],
  caption = "Altitude error at ground level, 2022-2024 (m)."
)
```

```{r}
#| fig-cap: "Accuracy: mean error per station."
fig("A_map_bias.png")
```

```{r}
#| fig-cap: "Precision: temporal SD of the error per station."
fig("A_map_sd.png")
```

```{r}
#| fig-cap: "Share of stations below a given |bias|, ERA5 single-levels vs ERA5-Land."
#| out-width: 70%
fig("A_sl_vs_land.png")
```

## By terrain, elevation, latitude and climate

```{r}
knitr::kable(
  A[group != "all" & dataset == "sl", .(group, subset, stations = n_stations,
    `median |bias|` = round(abs_bias_median, 1), `p90 |bias|` = round(abs_bias_p90, 1),
    `median SD` = round(sd_median, 1), `p90 SD` = round(sd_p90, 1), MAE = round(mae, 1),
    `p95 |err|` = round(p95_abs, 1))],
  caption = "ERA5 single-levels, by station class (m). Terrain = sub-grid orography SD in the ERA5 cell."
)
```

## Drivers

```{r}
#| fig-cap: "Left: bias against the gap between station elevation and ERA5 orography. Right: precision against sub-grid terrain roughness."
fig("A_drivers.png")
```

```{r}
d <- tab("A_drivers_station.csv")
knitr::kable(d[, .(response, term, `deviance explained (full)` = round(dev_expl_full, 3),
  `lost when dropped` = round(dev_expl_drop, 3), n)], caption = "Station-level drivers (GAM).")
d <- tab("A_drivers_obs.csv")
knitr::kable(d[, .(term, `deviance explained (full)` = round(dev_expl_full, 3),
  `lost when dropped` = round(dev_expl_drop, 3), n)],
  caption = "Observation-level drivers of the de-biased absolute error (GAM with station random effect).")
```

```{r}
#| fig-cap: "Diurnal and seasonal cycle of the de-biased error (RMS), by climate zone."
fig("A_cycles.png")
```

```{r}
#| fig-cap: "Change over time, for stations reporting in every year."
#| out-width: 70%
if (file.exists("output/figures/A_era.png")) fig("A_era.png")
```

```{r}
#| results: asis
if (file.exists("output/tables/A_era.csv")) {
  e <- tab("A_era.csv")
  print(knitr::kable(e[, lapply(.SD, function(x) round(x, 1)), .SDcols = names(e)],
    caption = "Change over time for stations reporting in every year (m)."))
}
```

# Results: in flight (Tier B)

```{r}
knitr::kable(
  B[, .(method, `height above ground` = hbin, n = format(n, big.mark = ","),
    bias = round(bias, 1), SD = round(sd, 1), MAE = round(mae, 1), RMSE = round(rmse, 1),
    `p95 |err|` = round(p95_abs, 1))],
  caption = "Error relative to the surface level of the same sounding, by height (m)."
)
```

```{r}
#| fig-cap: "Bias and SD of the altitude error against height above ground, for ERA5 single-levels and ERA5-Land, with the height distribution of geolocator flight points (right)."
fig("B_height.png")
```

```{r}
#| fig-cap: "The height-dependent error is almost entirely the difference between the temperature profile the formula assumes and the one the sonde observed."
#| out-width: 70%
fig("B_temperature.png")
```

```{r}
#| fig-cap: "Height-dependent bias by climate zone and season (solid: summer half-year, dashed: winter half-year; southern hemisphere shifted by six months)."
#| out-width: 80%
fig("B_height_climate.png")
```

```{r}
bw <- tab("B_bird_weighted.csv")
knitr::kable(bw[, .(method, bias = round(bias, 1), MAE = round(mae, 1), RMSE = round(rmse, 1))],
  caption = "Error averaged over the height distribution of geolocator flight points (m).")
```

```{r}
d <- tab("B_height_daynight_season.csv")[method == "GeoPressureR (ERA5 single-levels)"]
knitr::kable(dcast(d, hbin ~ daynight + season, value.var = "bias", fun.aggregate = function(x)
  round(x, 1)), caption = "Bias by height, day/night and season (m).")
```

## Can the formula be improved?

**Fitted lapse rate.** With the 2 m temperature, the lapse rate that best fits the radiosondes is
`r sprintf("%.1f", L_t2m)` K/km; with the virtual temperature it is `r sprintf("%.1f", L_tv)` K/km.
Both are well determined: across the 20 station splits the fitted value varies by
±`r sprintf("%.1f", ff[temperature == "Tv", L_cv_sd])` K/km (SD). Weighting all height bins equally
instead of by bird flight heights gives `r sprintf("%.1f", ff[temperature == "T2m", L_equal_bins])`
and `r sprintf("%.1f", ff[temperature == "Tv", L_equal_bins])` K/km. Accounting for humidity
explicitly thus brings the lapse rate back towards the standard value: about a third of the gap
between −6.5 K/km and the fitted dry-air lapse rate is a humidity effect.

```{r}
knitr::kable(
  fcv[, .(formula = method, `L fitted (K/km)` = sprintf("%.2f ± %.2f", 1000 * L_mean, 1000 * L_sd),
    bias = sprintf("%.1f ± %.1f", bias_mean, bias_sd), MAE = sprintf("%.1f ± %.1f", mae_mean, mae_sd),
    RMSE = sprintf("%.1f ± %.1f", rmse_mean, rmse_sd),
    `MAE gain (worst split)` = sprintf("%.1f (%.1f)", mae_gain_mean, mae_gain_min))],
  caption = "Bird-weighted error on held-out stations (mean ± SD over 20 random 50/50 station splits; the lapse rate is fitted on the other half). MAE gain: reduction of MAE relative to the current formula on the same held-out stations (m)."
)
```

**Out-of-sample performance.** On held-out stations, the virtual temperature alone reduces the
bird-weighted bias from `r fc("current", "bias_mean")` m to `r fc("tv", "bias_mean")` m, and
together with the fitted lapse rate to `r fc("tv_lapse", "bias_mean")` m. The mean absolute error
falls from `r fc("current", "mae_mean")` m to `r fc("tv_lapse", "mae_mean")` m, and it improves in
every one of the 20 splits. Held-out and in-sample errors are practically identical, as expected for a
single parameter estimated from ~300 stations.

```{r}
#| fig-cap: "Bias and SD against height above ground for the current formula and the three alternatives (all stations; lapse rates fitted on all stations), with the height distribution of geolocator flight points (right)."
fig("B_formula.png")
```

```{r}
knitr::kable(
  dcast(fh[, .(code = factor(code, c("current", "tv", "lapse", "tv_lapse")), agl_mid, hbin,
    bias = round(bias, 1))], agl_mid + hbin ~ code, value.var = "bias")[, !"agl_mid"],
  col.names = c("height above ground", fbw$method[match(c("current", "tv", "lapse", "tv_lapse"),
    fbw$code)]),
  caption = "Bias relative to the surface level, by height (m)."
)
```

**Error by height.** The corrected formula removes the systematic underestimation up to about 3 km
(`r fhb("tv_lapse", "(1e+03,1.5e+03]")` m at 1–1.5 km and `r fhb("tv_lapse", "(2e+03,3e+03]")` m
at 2–3 km, compared with `r fhb("current", "(1e+03,1.5e+03]")` m and
`r fhb("current", "(2e+03,3e+03]")` m). Above that it slightly overestimates
(`r fhb("tv_lapse", "(5e+03,6e+03]")` m at 5–6 km), because the fit is dominated by the heights at
which birds fly most. **The scatter does not change**: the SD at each height is the same for all four
formulas, because a constant lapse rate cannot follow the day-to-day temperature profile.

```{r}
#| fig-cap: "Height-dependent bias by climate zone and season, current formula (left) and virtual temperature with the fitted lapse rate (right). Solid: summer half-year; dashed: winter half-year."
fig("B_formula_climate.png")
```

```{r}
knitr::kable(
  fgrp[, .(climate, season, stations = n_st, `L with T2m (K/km)` = round(L_t2m, 1),
    `L with Tv (K/km)` = round(L_tv, 1))],
  caption = "Lapse rate fitted separately for each climate zone and half-year."
)
```

**Regional differences remain.** Fitted separately, the lapse rate (with virtual temperature)
ranges from `r sprintf("%.1f", min(fgrp$L_tv))` K/km (arid summer: steeper than standard, as a deep, well-mixed
boundary layer approaches the dry adiabatic 9.8 K/km) to `r sprintf("%.1f", max(fgrp$L_tv))` K/km
(polar winter, with persistent surface inversions). Continental and polar winters stay
underestimated with the global value, and arid summers become overestimated. A single global lapse
rate removes the mean bias, not these regional and seasonal patterns.

**Ground level.** With the fitted lapse rate, ground-level altitude is practically unchanged:
pooled MAE `r sprintf("%.2f", fg$mae)` m → `r sprintf("%.2f", fg$mae_new)` m, median station bias
`r sprintf("%.2f", fg$abs_bias_median)` m → `r sprintf("%.2f", fg$abs_bias_median_new)` m, largest
change of a single observation `r sprintf("%.1f", fg$max_abs_change)` m
(`r fg$n_stations` Tier A reference stations; the virtual-temperature factor, not available for
Tier A, changes ground-level altitude by less than 1.2% of the station-to-orography gap).

# Discussion and guidance for users

```{r}
tr <- A[dataset == "sl"]
t <- function(sub, col) round(tr[subset == sub][[col]], 1)
e <- if (file.exists("output/tables/A_era.csv")) tab("A_era.csv") else NULL
te <- tab("B_temperature_explanation.csv")
```

## What users can expect

| Situation | Typical error | Driver |
|---|---|---|
| Absolute altitude on the ground, flat terrain | bias `r t("flat (<20 m)", "abs_bias_median")` m (p90 `r t("flat (<20 m)", "abs_bias_p90")` m), MAE `r t("flat (<20 m)", "mae")` m | ERA5 surface pressure at 0.25° |
| Absolute altitude on the ground, mountains (sub-grid SD > 300 m) | bias `r t("mountain (>300 m)", "abs_bias_median")` m (p90 `r t("mountain (>300 m)", "abs_bias_p90")` m), MAE `r t("mountain (>300 m)", "mae")` m | terrain the 0.25° grid cannot resolve |
| Altitude *changes* at one site (precision) | SD `r t("flat (<20 m)", "sd_median")` m (flat) to `r t("mountain (>300 m)", "sd_median")` m (mountains) | as above; almost no diurnal or seasonal cycle |
| In flight, 1–1.5 km above ground | `r h("B_rel_bias_1000_1500", 0)` m bias, RMSE `r h("B_rel_rmse_1000_1500", 0)` m (added to the ground-level error) | assumed temperature profile |
| In flight, 2–3 km above ground | `r h("B_rel_bias_2000_3000", 0)` m bias, RMSE `r h("B_rel_rmse_2000_3000", 0)` m | as above; ~2x in continental/polar winter |

**On the ground, the error is mostly a fixed offset per place, not noise.** Station identity
explains almost all of the explainable variation in the de-biased error; hour of day, season,
boundary layer height, surface inversions and pressure tendency each add less than 0.5% of
deviance. Precision (SD ~4 m) is therefore better than absolute accuracy (MAE ~7 m), and roughly
constant in time. What changes the error is *where*: the gap between the
true ground and ERA5's smoothed orography, and sub-grid terrain roughness.

**In flight, the error grows with height and is systematic.** The barometric formula extrapolates
the ERA5 2 m temperature upwards with a standard lapse rate of −6.5 K/km and ignores humidity. The
real atmosphere is usually warmer than that (shallower lapse rates, surface inversions, and the
virtual-temperature effect of water vapour), so the layer is too cold in the formula and the bird is
placed too low. The error predicted from the sonde's own temperature profile explains
`r round(100 * te$r2)`% of the variance (r = `r round(te$r, 2)`, slope `r round(te$slope, 2)`); the
remaining offset of `r round(te$intercept)` m is consistent with humidity, which the dry-bulb
comparison leaves out (the virtual-temperature correction alone removes
`r sprintf("%.1f", fbw[code == "current", bias] - fbw[code == "tv", bias])` m of the bird-weighted
bias). It is worst over cold continental and polar surfaces in winter, where strong
inversions sit under much warmer air, and in the humid tropics.

**Flight altitudes from GeoPressureR are therefore conservative (slightly too low).** For a bird at
2 km above the ground, expect it about 25–35 m too low, with a scatter (SD) of 30–50 m; roughly
double that in a continental or polar winter.

**ERA5 has improved.** For the same stations, the median ground-level bias fell from
`r if (!is.null(e)) round(e[year == min(year), abs_bias_median], 1) else "NA"` m in
`r if (!is.null(e)) min(e$year) else "NA"` to
`r if (!is.null(e)) round(e[year == max(year), abs_bias_median], 1) else "NA"` m in
`r if (!is.null(e)) max(e$year) else "NA"`, and the precision improved by about a third. Tracks
from the 1990s and earlier should be expected to be somewhat less accurate.

**ERA5-Land vs ERA5 single-levels.** ERA5-Land's median ground-level bias is about four times
larger. Its height-*relative* errors in flight are similar to single-levels, because the error
relative to the surface level cancels its static orography inconsistency. This confirms that its
problem is a fixed spatial offset
([GeoPressureAPI#27](https://github.com/GeoPressure/GeoPressureAPI/issues/27)), which matters
for absolute altitude and for comparing altitudes between places.

## Improving the formula

**Recommendation for GeoPressureR.** The virtual temperature is a correction of a simplification in
the current formula, has no free parameter, and removes about 40% of the in-flight bias on its
own; it only requires reading the ERA5 2 m dewpoint in addition to surface pressure and 2 m
temperature. Combined with a lapse rate of `r sprintf("%.1f", L_tv)` K/km, the in-flight bias
averaged over bird flight heights drops from `r fc("current", "bias_mean")` m to
`r fc("tv_lapse", "bias_mean")` m and the MAE from `r fc("current", "mae_mean")` m to
`r fc("tv_lapse", "mae_mean")` m, on stations not used for the fit, with no effect on ground-level
altitude. The lapse rate is an empirical estimate, but a robust one: it is stable across station
splits and objectives, and it lies in the range reported for the lower troposphere near the
surface. Fitting the lapse rate alone (with the dry 2 m temperature, `r sprintf("%.1f", L_t2m)` K/km)
performs about as well statistically. We nevertheless recommend the combination: the virtual
temperature is required by the physics, whereas a lapse rate fitted without it mixes the humidity
and temperature-profile effects into one constant with no physical meaning, which is harder to
interpret and to refine (e.g. into a lapse rate varying with location and season).

**What a constant lapse rate cannot fix.** The scatter in flight, and the remaining regional and
seasonal biases (continental and polar winters too low, arid summers too high), come from
departures of the actual temperature profile from any fixed one, mainly surface inversions and
deep convective boundary layers. Two further steps could address them: a lapse rate that varies with
location and season (e.g. the climate-zone values above, or a climatology derived from ERA5), or
using the actual ERA5 temperature profile on pressure levels, or ERA5 geopotential at the bird's
pressure directly. The latter would remove most of the temperature-driven variance, which explains
`r round(100 * te$r2)`% of the in-flight error, but it needs far more ERA5 data per point and is
left for future evaluation.

## Limitations

- **Not independent of ERA5.** Station surface pressure and radiosondes are both assimilated by
  ERA5, so errors *at these sites* are optimistic. Remote areas without observations are likely
  somewhat worse, especially for ground-level bias.
- **The sensor is not included.** Geolocator pressure sensors add their own offset, drift and
  resolution (1 hPa ≈ 8 m near the ground). The numbers here are the error of the retrieval method
  given a perfect pressure reading.
- **Radiosonde geometry.** The ERA5 column at the launch site is used; balloon drift (typically a few
  km below 5 km) is ignored. Sonde heights are themselves derived hypsometrically. The ~10 m scatter
  within the first 100 m partly reflects integer-metre reporting and interpolated low levels.
- **Reference screen.** `r hi("A_n_excluded")` surface stations were excluded for an implausible
  offset or a step change. Neighbours and DEMs support this where available, but not every excluded
  station could be checked independently. Conversely, the reference set still contains undetected
  small reference errors, so its accuracy is an upper bound on ERA5's.
- **Sampling.** Stations are thinned for spatial balance, but coverage is still sparse in parts of
  Africa, South America and over the oceans (Tier A is land-only).

# Data and code availability

Code: <https://github.com/GeoPressure/altitude-validation>. ERA5 and ERA5-Land: Copernicus Climate
Change Service (Hersbach et al. 2020; Muñoz-Sabater et al. 2021), via the ECMWF ARCO archive.
HadISD: Met Office Hadley Centre, Non-Commercial Government Licence; this product may contain data
governed by WMO Resolution 40 Annex 1. IGRA2: NOAA NCEI. Bird heights: GeoLocator master data
package (<https://doi.org/10.5281/zenodo.18187092>).

# Appendix: reference screen {#sec-appendix}

## Sensitivity to the screen

```{r}
sens <- tab("A_sensitivity.csv")
knitr::kable(sens[, .(`offset rule` = rule, `step-change stations` = steps, stations = n_stations,
  `median |bias|` = round(abs_bias_median, 1), `p90 |bias|` = round(abs_bias_p90, 1),
  `median SD` = round(sd_median, 1), MAE = round(mae, 1), RMSE = round(rmse, 1),
  `p95 |err|` = round(p95_abs, 1))],
  caption = "Ground-level statistics (ERA5 single-levels, 2022-2024, m) for looser and stricter plausibility bounds (|bias| <= a + b x |station elevation - ERA5 orography|), with the step-change stations kept or excluded. The row used in the report is '30 m + 10% (used)' with step-change stations excluded.")
```

The median bias and median SD are insensitive to the screen. MAE, RMSE and the 95th percentile are
dominated by the few stations with offsets of hundreds of metres when those are included. Loosening
the bound to 60 m + 20% adds back `r sens[rule == "60 m + 20%" & steps == "excluded", n_stations] - sens[rule == "30 m + 10% (used)" & steps == "excluded", n_stations]`
stations and raises the MAE from `r round(sens[rule == "30 m + 10% (used)" & steps == "excluded", mae], 1)`
to `r round(sens[rule == "60 m + 20%" & steps == "excluded", mae], 1)` m.

## Basis of the plausibility bound

```{r}
knitr::kable(bf[, lapply(.SD, function(x) round(x, 1))],
  caption = "|bias| where ERA5 needs no extrapolation: flat terrain (sub-grid SD < 20 m) and station within 30 m of ERA5 orography (m).")
knitr::kable(bg[, .(`height above ground (m)` = hbin, stations = n_stations,
  `median (% of height)` = round(median_pct, 2), `99th pct of |.| (%)` = round(p99_abs_pct, 1),
  `max |.| (%)` = round(max_abs_pct, 1))],
  caption = "Extrapolation error over a height gap, measured by radiosondes: per-station mean error relative to the surface level, divided by the height above ground.")
```

## Excluded stations

```{r}
options(knitr.kable.NA = "")
ex[!grepl("step", reason), `:=`(step = NA, step_at = NA)]
knitr::kable(ex[, .(station = name, reason, `elev.` = elev, `ERA5 orog.` = z_sl, bias, SD = sd,
  step, `step from` = step_at, SRTM = dem_srtm, ASTER = dem_aster,
  `DEM refutes elev.` = dem_refutes, `neighbour bias` = nb_bias_median,
  `neighbour km` = nb_km_min)],
  caption = "Stations excluded from the reference set (m). SRTM/ASTER: DEM at the listed coordinates; 'DEM refutes' = listed elevation more than 15 m outside the DEM range over the coordinate-uncertainty box in every available DEM. Neighbour bias: median ERA5 bias in 2023 at up to two HadISD stations within 50 km (blank: none, or not tested).")
```

# References

- Bolton D (1980) The computation of equivalent potential temperature. *Monthly Weather Review*
  108:1046–1053. <https://doi.org/10.1175/1520-0493(1980)108%3C1046:TCOEPT%3E2.0.CO;2>
- Dunn RJH, Willett KM, Thorne PW, et al. (2012) HadISD: a quality-controlled global synoptic report
  database for selected variables at long-term stations from 1973–2011. *Climate of the Past*
  8:1649–1679. <https://doi.org/10.5194/cp-8-1649-2012>
- Dunn RJH, Willett KM, Parker DE, Mitchell L (2016) Expanding HadISD: quality-controlled, sub-daily
  station data from 1931. *Geoscientific Instrumentation, Methods and Data Systems* 5:473–491.
  <https://doi.org/10.5194/gi-5-473-2016>
- Durre I, Vose RS, Wuertz DB (2006) Overview of the Integrated Global Radiosonde Archive. *Journal
  of Climate* 19:53–68. <https://doi.org/10.1175/JCLI3594.1>
- Durre I, Yin X, Vose RS, et al. (2018) Enhancing the data coverage in the Integrated Global
  Radiosonde Archive. *Journal of Atmospheric and Oceanic Technology* 35:1753–1770.
  <https://doi.org/10.1175/JTECH-D-17-0223.1>
- ECMWF (2025) Guidance on SYNOP surface pressure monitoring issues. ECMWF Confluence (E. Kuscu).
  <https://confluence.ecmwf.int/pages/viewpage.action?pageId=298952900>
- Hersbach H, Bell B, Berrisford P, et al. (2020) The ERA5 global reanalysis. *Quarterly Journal of
  the Royal Meteorological Society* 146:1999–2049. <https://doi.org/10.1002/qj.3803>
- ICAO (1993) *Manual of the ICAO Standard Atmosphere*, Doc 7488/3, 3rd ed. International Civil
  Aviation Organization, Montreal.
- ISO (1975) *ISO 2533:1975 Standard Atmosphere*. International Organization for Standardization.
- Minder JR, Mote PW, Lundquist JD (2010) Surface temperature lapse rates over complex terrain:
  lessons from the Cascade Mountains. *Journal of Geophysical Research* 115:D14122.
  <https://doi.org/10.1029/2009JD013493>
- Muñoz-Sabater J, Dutra E, Agustí-Panareda A, et al. (2021) ERA5-Land: a state-of-the-art global
  reanalysis dataset for land applications. *Earth System Science Data* 13:4349–4383.
  <https://doi.org/10.5194/essd-13-4349-2021>
- NASA/METI/AIST/Japan Spacesystems and U.S./Japan ASTER Science Team (2019) ASTER Global Digital
  Elevation Model V003. NASA EOSDIS Land Processes DAAC. <https://doi.org/10.5067/ASTER/ASTGTM.003>
- NASA JPL (2013) NASA Shuttle Radar Topography Mission Global 1 arc second. NASA EOSDIS Land
  Processes DAAC. <https://doi.org/10.5067/MEaSUREs/SRTM/SRTMGL1.003>
- Rodríguez E, Morris CS, Belz JE (2006) A global assessment of the SRTM performance.
  *Photogrammetric Engineering & Remote Sensing* 72:249–260. <https://doi.org/10.14358/PERS.72.3.249>
- Rolland C (2003) Spatial and seasonal variations of air temperature lapse rates in Alpine regions.
  *Journal of Climate* 16:1032–1046.
  <https://doi.org/10.1175/1520-0442(2003)016%3C1032:SASVOA%3E2.0.CO;2>
- Smith A, Lott N, Vose R (2011) The Integrated Surface Database: recent developments and
  partnerships. *Bulletin of the American Meteorological Society* 92:704–708.
  <https://doi.org/10.1175/2011BAMS3015.1>
- Stone PH, Carlson JH (1979) Atmospheric lapse rate regimes and their parameterization. *Journal
  of the Atmospheric Sciences* 36:415–423.
  <https://doi.org/10.1175/1520-0469(1979)036%3C0415:ALRRAT%3E2.0.CO;2>
- Vasiljevic D, Andersson E, Isaksen L, Garcia-Mendez A (2006) Surface pressure bias correction in
  data assimilation. *ECMWF Newsletter* 108:20–27. <https://doi.org/10.21957/uv295rfmx5>
- Wallace JM, Hobbs PV (2006) *Atmospheric Science: An Introductory Survey*, 2nd ed., ch. 3.
  Academic Press. <https://doi.org/10.1016/B978-0-12-732951-2.50008-9>
- WMO (1968) *Methods in Use for the Reduction of Atmospheric Pressure*. Technical Note 91,
  WMO-No. 226. World Meteorological Organization, Geneva.
- WMO (2023) *Guide to Instruments and Methods of Observation*, Volume I (WMO-No. 8), chapters 1
  (station elevation) and 3 (atmospheric pressure). World Meteorological Organization, Geneva.
